---
output:
  pdf_document:
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{amsmath}
  - \input{../../_common/preamble.tex}
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

```{=latex}
\makecoverpage{Chapter 12 --- Unsupervised Learning \\ Exercises 4 and 11}
```

```{r libraries}
library(ISLR2)
```

# Exercise 4: Single Linkage versus Complete Linkage

Hierarchical clustering starts with every observation in its own cluster and
then repeatedly merges the two clusters that are closest according to a
chosen linkage rule. The linkage rule does not change the pairwise distances
between individual observations. It changes how we define the distance
between two *clusters*.

For two clusters $A$ and $B$, single linkage uses the closest pair of points:

$$
D_{\mathrm{single}}(A,B) = \min_{i\in A,\,j\in B} d(i,j).
$$

Complete linkage uses the farthest pair of points:

$$
D_{\mathrm{complete}}(A,B) = \max_{i\in A,\,j\in B} d(i,j).
$$

Because a maximum is always at least as large as a minimum,

$$
D_{\mathrm{complete}}(A,B) \geq D_{\mathrm{single}}(A,B)
$$

for every pair of clusters $A$ and $B$. This simple inequality is the key to
both parts of the exercise.

## (a) The clusters $\{1,2,3\}$ and $\{4,5\}$

The two methods merge the same two clusters at some point in their respective
dendrograms. The question is whether the fusion occurs higher, at the same
height, or whether there is not enough information to tell.

The answer is that the complete-linkage fusion occurs **higher than or at the
same height as** the single-linkage fusion. It cannot occur lower.

The reason is direct. For these particular clusters, write

$$
A=\{1,2,3\}, \qquad B=\{4,5\}.
$$

Then

$$
D_{\mathrm{single}}(A,B)
= \min\{d(1,4),d(1,5),d(2,4),d(2,5),d(3,4),d(3,5)\},
$$

whereas

$$
D_{\mathrm{complete}}(A,B)
= \max\{d(1,4),d(1,5),d(2,4),d(2,5),d(3,4),d(3,5)\}.
$$

The maximum is at least the minimum. Since dendrogram height records the
linkage distance at which a fusion occurs, the complete-linkage height is at
least the single-linkage height.

It is important to distinguish this conclusion from the stronger statement
that the heights must be different. They need not be. If all six cross-cluster
distances happen to be equal, then their minimum and maximum are equal and
both methods fuse the clusters at the same height. Without the actual
pairwise distances, we cannot decide whether the inequality is strict.

Therefore, the precise answer is:

$$
\boxed{
\text{Complete linkage fuses }\{1,2,3\}\text{ and }\{4,5\}
\text{ at a height at least as large as single linkage.}
}
$$

## (b) The clusters $\{5\}$ and $\{6\}$

Here each cluster contains only one observation:

$$
A=\{5\}, \qquad B=\{6\}.
$$

There is only one cross-cluster distance, $d(5,6)$. Therefore the minimum
and maximum over the set containing this one value are identical:

$$
D_{\mathrm{single}}(\{5\},\{6\})
= d(5,6)
= D_{\mathrm{complete}}(\{5\},\{6\}).
$$

Consequently, the two methods fuse observations 5 and 6 at **exactly the same
height**.

This conclusion does not depend on the locations of observations 5 and 6 or
on the rest of the dataset. Whenever both clusters are singletons, every
linkage definition based on the minimum or maximum of cross-cluster distances
reduces to the same one pairwise distance.

Thus,

$$
\boxed{
\text{The single-linkage and complete-linkage fusions occur at the same height.}
}
$$

# Exercise 11: Matrix Completion

Matrix completion estimates entries that are missing from a matrix by using
low-dimensional structure in the entries that are observed. The central
assumption is that the complete matrix is approximately low rank. In other
words, although the matrix may have many rows and columns, most of its
variation can be represented using a relatively small number of latent
components.

Suppose $X$ is the unknown complete matrix and $\Omega$ is the set of observed
entries. We observe $X_{ij}$ only for $(i,j)\in\Omega$. A rank-$M$
approximation seeks a matrix $\widehat X$ of the form

$$
\widehat X_M = U_M D_M V_M^T,
$$

where $U_M$ and $V_M$ contain the first $M$ singular vectors and $D_M$
contains the first $M$ singular values. Equivalently, the singular value
decomposition gives

$$
X \approx \sum_{m=1}^{M} d_m u_m v_m^T.
$$

Each outer product $u_m v_m^T$ is a rank-one pattern. Increasing $M$ allows
more patterns to be represented, so approximation bias generally decreases;
however, larger values of $M$ can also fit noise and are more expensive to
estimate when many entries are missing.

## Algorithm 12.1 in plain language

The matrix-completion algorithm alternates between two operations:

1. fill the missing entries with current guesses;
2. compute a rank-$M$ approximation of the completed matrix using the SVD,
   then use that approximation to update the missing entries.

The observed entries are never changed. They anchor the procedure to the
actual data, while the unobserved entries are repeatedly replaced by values
consistent with the estimated low-rank structure.

Let $P_\Omega$ be the projection that keeps observed entries and sets missing
entries to zero. The update can be written conceptually as

$$
X^{(t+1)}
= P_\Omega(X) + P_{\Omega^c}\left(\mathcal{P}_M(X^{(t)})\right),
$$

where $\mathcal{P}_M$ means the best rank-$M$ SVD approximation and
$\Omega^c$ is the set of missing entries. The first term restores the observed
values; the second supplies estimates only where values are missing.

The function below also tracks a convergence quantity. After each update, it
computes the relative change between consecutive completed matrices:

$$
\delta_t =
\frac{\left\|X^{(t+1)}-X^{(t)}\right\|_F}
{\max\left(\left\|X^{(t)}\right\|_F,\varepsilon\right)},
$$

where $\|\cdot\|_F$ is the Frobenius norm and $\varepsilon$ prevents division
by zero. Iteration stops when $\delta_t$ is below `tol` or when `max_iter`
is reached.

```{r matrix-completion-function}
rank_m_approximation <- function(x, M) {
  decomposition <- svd(x, nu = M, nv = M)
  rank_used <- min(M, length(decomposition$d))

  decomposition$u[, seq_len(rank_used), drop = FALSE] %*%
    (decomposition$d[seq_len(rank_used)] *
       t(decomposition$v[, seq_len(rank_used), drop = FALSE]))
}

matrix_completion <- function(X_observed,
                              observed,
                              M,
                              tol = 1e-5,
                              max_iter = 1000,
                              verbose = FALSE) {
  stopifnot(is.matrix(X_observed), is.matrix(observed))
  stopifnot(all(dim(X_observed) == dim(observed)))
  stopifnot(length(M) == 1, M >= 1, M == as.integer(M))
  stopifnot(length(tol) == 1, tol > 0)
  stopifnot(length(max_iter) == 1, max_iter >= 1)

  if (!all(is.finite(X_observed[observed]))) {
    stop("Observed entries must be finite.")
  }
  if (!any(observed)) {
    stop("At least one entry must be observed.")
  }

  # Column means provide a stable starting value for each missing entry.
  completed <- X_observed
  for (j in seq_len(ncol(completed))) {
    observed_in_column <- observed[, j]
    column_mean <- if (any(observed_in_column)) {
      mean(X_observed[observed_in_column, j])
    } else {
      mean(X_observed[observed], na.rm = TRUE)
    }
    completed[!observed_in_column, j] <- column_mean
  }
  completed[observed] <- X_observed[observed]

  convergence <- numeric(max_iter)
  approximation <- completed
  iteration <- 0L

  for (step in seq_len(max_iter)) {
    approximation <- rank_m_approximation(completed, M)
    updated <- completed
    updated[!observed] <- approximation[!observed]
    updated[observed] <- X_observed[observed]

    relative_change <- sqrt(sum((updated - completed)^2)) /
      max(sqrt(sum(completed^2)), .Machine$double.eps)
    convergence[step] <- relative_change
    iteration <- step

    if (verbose) {
      cat(sprintf("iteration %d: relative change = %.7g\n",
                  step, relative_change))
    }

    completed <- updated
    if (relative_change < tol) {
      break
    }
  }

  list(
    completed = completed,
    iterations = iteration,
    convergence = convergence[seq_len(iteration)],
    converged = convergence[iteration] < tol,
    rank = M
  )
}
```

The observed-entry constraint is worth checking explicitly. Even after many
iterations, the returned matrix must reproduce every value that was supplied
to the function:

```{r matrix-completion-check}
set.seed(12)
toy <- matrix(rnorm(30), nrow = 6)
toy_observed <- matrix(runif(length(toy)) > 0.25,
                       nrow = nrow(toy), ncol = ncol(toy))
toy_input <- toy
toy_input[!toy_observed] <- NA_real_

toy_fit <- matrix_completion(toy_input, toy_observed, M = 2)

max_observed_difference <- max(abs(
  toy_fit$completed[toy_observed] - toy[toy_observed]
))
max_observed_difference
```

The maximum difference should be zero up to floating-point precision. This
check matters because a method that changes observed entries would no longer
be completing the matrix; it would simply be fitting a low-rank approximation
to a modified dataset.

## Preparing the `Boston` experiment

The `Boston` dataset contains 506 observations and 13 numeric features. The
exercise asks us to standardize the features first. For every feature $j$, the
transformation is

$$
Z_{ij} = \frac{X_{ij}-\bar X_j}{s_j},
$$

so each column has mean approximately zero and standard deviation one. This
is important because the SVD is scale-sensitive: without standardization, a
feature measured in large units could dominate the decomposition simply
because of its units rather than because it carries more structure.

```{r boston-standardize}
Boston_features <- as.matrix(ISLR2::Boston[, setdiff(names(ISLR2::Boston), "medv")])
Boston_scaled <- scale(Boston_features)

c(
  largest_absolute_column_mean = max(abs(colMeans(Boston_scaled))),
  largest_standard_deviation_error = max(abs(apply(Boston_scaled, 2, sd) - 1))
)
```

We evaluate approximation error only on entries that were deliberately hidden.
Because the original `Boston_scaled` matrix is available to us in this
experiment, we can compare an imputed value with its known truth. For a mask
$H$ of hidden entries, the relative error is

$$
\operatorname{RelErr}(\widehat X)
= \frac{\|H\odot(\widehat X-X)\|_F}
{\|H\odot X\|_F},
$$

where $\odot$ denotes elementwise multiplication. Evaluating on hidden
entries is the matrix-completion analogue of using a test set: those values
were not supplied to the algorithm.

## One worked run

This first run hides 15% of the entries, uses a rank-$M=3$ approximation, and
prints the convergence progress. The hidden entries are created from a random
permutation, so the missingness is random while still allowing us to construct
nested masks for the full experiment below.

```{r matrix-completion-worked-run}
relative_error <- function(estimate, truth, hidden) {
  sqrt(sum((estimate[hidden] - truth[hidden])^2)) /
    max(sqrt(sum(truth[hidden]^2)), .Machine$double.eps)
}

set.seed(11)
missing_fraction <- 0.15
hidden_count <- ceiling(missing_fraction * length(Boston_scaled))
hidden_indices <- sample(seq_along(Boston_scaled), hidden_count)
hidden <- matrix(FALSE, nrow(Boston_scaled), ncol(Boston_scaled))
hidden[hidden_indices] <- TRUE
observed <- !hidden

input <- Boston_scaled
input[hidden] <- NA_real_
worked_run <- matrix_completion(input, observed, M = 3,
                                tol = 1e-5, max_iter = 500,
                                verbose = TRUE)

list(
  iterations = worked_run$iterations,
  converged = worked_run$converged,
  final_relative_change = tail(worked_run$convergence, 1),
  hidden_entry_relative_error = relative_error(
    worked_run$completed, Boston_scaled, hidden
  )
)
```

The convergence value and the approximation error answer different questions.
The convergence value asks whether successive imputations have stopped
changing. The hidden-entry error asks whether the final imputations are close
to the values that were actually hidden. Convergence is necessary for a stable
iteration, but it does not by itself guarantee a small prediction error.

## Nested missingness experiment

We now follow the exercise exactly. The missing fractions are 5%, 10%, ..., 30%.
For each of 10 repetitions, one random ordering of all matrix entries is
created. The first 5% of that ordering are hidden for the 5% experiment, the
first 10% are hidden for the 10% experiment, and so on. Thus the missing sets
are nested within each repetition.

For every missingness level and every $M\in\{1,2,\ldots,8\}$, we run the
completion algorithm and calculate the hidden-entry relative error.

```{r matrix-completion-experiment, cache=TRUE}
set.seed(2026)
missing_fractions <- seq(0.05, 0.30, by = 0.05)
rank_values <- 1:8
repetitions <- 10

experiment_results <- vector("list", length(missing_fractions) *
                               length(rank_values) * repetitions)
result_index <- 1L

for (replication in seq_len(repetitions)) {
  permutation <- sample(seq_along(Boston_scaled))

  for (missing_fraction in missing_fractions) {
    hidden_count <- ceiling(missing_fraction * length(Boston_scaled))
    hidden <- matrix(FALSE, nrow(Boston_scaled), ncol(Boston_scaled))
    hidden[permutation[seq_len(hidden_count)]] <- TRUE
    observed <- !hidden

    input <- Boston_scaled
    input[hidden] <- NA_real_

    for (M in rank_values) {
      fit <- matrix_completion(input, observed, M = M,
                               tol = 1e-5, max_iter = 500)

      experiment_results[[result_index]] <- data.frame(
        repetition = replication,
        missing_fraction = missing_fraction,
        M = M,
        approximation_error = relative_error(
          fit$completed, Boston_scaled, hidden
        ),
        iterations = fit$iterations,
        converged = fit$converged
      )
      result_index <- result_index + 1L
    }
  }
}

experiment_results <- do.call(rbind, experiment_results)
head(experiment_results)
```

## Average approximation error

```{r matrix-completion-summary}
error_summary <- aggregate(
  approximation_error ~ missing_fraction + M,
  data = experiment_results,
  FUN = mean
)
error_summary
```

```{r matrix-completion-plot, fig.width=7, fig.height=5}
plot(NA, xlim = range(missing_fractions),
     ylim = range(error_summary$approximation_error),
     xlab = "Fraction of observations missing",
     ylab = "Mean relative approximation error",
     main = "Matrix-completion error across missingness and rank")

for (M in rank_values) {
  curve_data <- error_summary[error_summary$M == M, ]
  lines(curve_data$missing_fraction, curve_data$approximation_error,
        type = "b", pch = 19, col = M)
}
legend("topleft", legend = paste("M =", rank_values),
       col = rank_values, lty = 1, pch = 19, ncol = 2, bty = "n")
```

The expected pattern is that error increases as the missing fraction grows:
with more entries hidden, the algorithm has less information from which to
recover the latent structure. Increasing $M$ often lowers error at first
because the approximation is more flexible. However, the benefit can flatten
or reverse when a high-rank approximation starts fitting idiosyncratic noise
rather than the stable structure shared across the matrix.

The exact ranking of the curves is an empirical result, not a theorem. It
depends on the random masks, the tolerance, and the structure of the data. The
10 repetitions reduce the chance that one unusually easy or difficult mask
determines the conclusion.

## Convergence diagnostics

The function was also required to track iteration count. We summarize the
average number of iterations and the fraction of fits that reached the
specified tolerance.

```{r matrix-completion-convergence}
iteration_summary <- aggregate(
  iterations ~ missing_fraction + M,
  data = experiment_results,
  FUN = mean
)

convergence_summary <- aggregate(
  converged ~ missing_fraction + M,
  data = experiment_results,
  FUN = mean
)

list(
  mean_iterations = iteration_summary,
  fraction_converged = convergence_summary
)
```

A larger missing fraction can require more iterations because the observed
entries provide a weaker anchor for the missing values. If a fit reaches
`max_iter` before the tolerance is met, that is not automatically a failure:
it is a diagnostic signal that the chosen tolerance or rank may require more
iterations, or that the alternating SVD updates are converging slowly.

# Conclusion

Exercise 4 illustrates how a precise definition of cluster distance controls
the geometry of a dendrogram. Complete linkage cannot assign a lower distance
to a pair of clusters than single linkage because a maximum cannot be below a
minimum. For singleton clusters, both definitions reduce to the same pairwise
distance.

Exercise 11 turns the low-rank idea into an algorithm. Missing entries are
initialized, a rank-$M$ SVD approximation is computed, and only the missing
entries are replaced. The observed entries remain fixed throughout. The
experiment then evaluates the method honestly on entries that were hidden from
it, across nested missingness levels, ranks $M=1$ through $8$, and 10 random
repetitions. This separates algorithmic convergence from predictive
approximation quality and makes the effect of missingness visible.
