---
output:
  pdf_document:
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{amsmath}
  - \input{../../_common/preamble.tex}
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

```{=latex}
\makecoverpage{Chapter 12 --- Unsupervised Learning \\ Lab}
```

```{r libraries}
library(ISLR2)
```

# Introduction

Supervised learning has a response variable that tells us what we are trying
to predict. Unsupervised learning starts without that response. Instead, it
looks for useful structure in the predictors themselves.

This lab develops the main tools from Chapter 12 of *An Introduction to
Statistical Learning*:

1. principal component analysis (PCA), which replaces many correlated
   predictors with a smaller number of directions that explain their shared
   variation;
2. hierarchical clustering, which builds a tree of increasingly large
   clusters; and
3. $K$-means clustering, which directly searches for a chosen number of
   compact groups.

The methods answer related but different questions. PCA is primarily a
continuous dimension-reduction method. Clustering is primarily a grouping
method. Neither method automatically discovers the one true answer: the
analyst must choose scales, distances, tuning values, and interpretations
carefully.

# Principal Component Analysis

## Why dimension reduction is useful

Suppose a dataset contains $p$ quantitative variables. Plotting or modelling
all $p$ variables can be difficult, especially when several variables measure
similar underlying phenomena. PCA constructs new variables called principal
components. Each component is a linear combination of the original variables:

$$
Z_m = \phi_{1m}X_1 + \phi_{2m}X_2 + \cdots + \phi_{pm}X_p.
$$

The loading vector $\phi_m=(\phi_{1m},\ldots,\phi_{pm})$ determines the
 direction of the component. PCA chooses the first direction to maximise the
variance of the projected observations, subject to $\|\phi_1\|=1$:

$$
\phi_1 = \arg\max_{\|\phi\|=1}
\operatorname{Var}(X\phi).
$$

The second component is the direction with the largest remaining variance,
subject to being orthogonal to the first, and so on. The components are
uncorrelated by construction and are ordered from greatest to least variance.

## PCA with the `USArrests` data

`USArrests` contains four crime-related measurements for the 50 US states.
The variables are measured on different scales, so the analysis should
standardize them first. Otherwise, a variable with larger numerical units
would influence the components more heavily even if it were not more
informative.

```{r pca-usarrests-prepare}
data(USArrests)
summary(USArrests)

USArrests_scaled <- scale(USArrests)
apply(USArrests_scaled, 2, mean)
apply(USArrests_scaled, 2, sd)
```

The sample covariance matrix shows how the original variables move together.
PCA diagonalizes this covariance matrix: its eigenvectors give the loading
vectors, and its eigenvalues give the variances of the corresponding
components.

```{r pca-usarrests}
usarrests.pca <- prcomp(USArrests, scale. = TRUE)
summary(usarrests.pca)
```

The standard deviations reported by `prcomp()` are the square roots of the
component variances. The proportion of variance explained by component $m$ is
therefore

$$
\operatorname{PVE}_m =
\frac{d_m^2}{\sum_{j=1}^{p}d_j^2},
$$

where $d_m$ is the reported standard deviation of component $m$. The
cumulative PVE tells us how many components are needed to retain a chosen
amount of total variation.

```{r pca-usarrests-table}
usarrests_pve <- data.frame(
  Component = seq_along(usarrests.pca$sdev),
  Standard_deviation = usarrests.pca$sdev,
  Proportion_variance = usarrests.pca$sdev^2 /
    sum(usarrests.pca$sdev^2),
  Cumulative_variance = cumsum(usarrests.pca$sdev^2) /
    sum(usarrests.pca$sdev^2)
)
usarrests_pve
```

```{r pca-usarrests-scree, fig.width=6, fig.height=4}
plot(usarrests_pve$Component, usarrests_pve$Proportion_variance,
     type = "b", pch = 19,
     xlab = "Principal component",
     ylab = "Proportion of variance explained",
     main = "USArrests: scree plot")
```

The loadings show which original variables contribute to each direction. A
large absolute loading means that the variable is important for that
component; the sign is a direction convention, because multiplying an entire
component by $-1$ gives the same underlying axis.

```{r pca-usarrests-loadings}
round(usarrests.pca$rotation, 3)
```

The first component often places violent and property crime measures in a
similar direction, while the second contrasts the variables that behave
differently. This interpretation should be made from the loading pattern,
not from the component name alone.

## Visualising observations in component space

```{r pca-usarrests-biplot, fig.width=7, fig.height=6}
biplot(usarrests.pca, scale = 0, cex = 0.7,
       main = "USArrests represented in the first two PCs")
```

A biplot displays both observations and variable directions. States that are
near one another have similar standardized crime profiles. Variables pointing
in similar directions are positively associated; variables pointing in
opposite directions are negatively associated in the displayed component
plane.

# Hierarchical Clustering

## Distances and linkage

Clustering begins with a measure of dissimilarity between observations. With
standardized quantitative variables, Euclidean distance between observations
$i$ and $j$ is

$$
 d(i,j) = \sqrt{\sum_{k=1}^{p}(x_{ik}-x_{jk})^2}.
$$

Hierarchical clustering initially treats every observation as its own
cluster. At each step it merges the two closest clusters and records the
fusion distance. The complete sequence of merges is represented by a
dendrogram.

Different linkage rules define cluster distance differently. For clusters $A$
and $B$:

$$
D_{\mathrm{single}}(A,B) = \min_{i\in A,j\in B}d(i,j),
$$

$$
D_{\mathrm{complete}}(A,B) = \max_{i\in A,j\in B}d(i,j),
$$

and average linkage uses the mean of all cross-cluster distances. Single
linkage can join groups through chains of nearby observations, while complete
linkage tends to favour compact groups.

## `USArrests` dendrograms

We use the standardized data so that the distance is not dominated by the
measurement units of a single variable.

```{r hclust-usarrests}
distance.usarrests <- dist(USArrests_scaled)

hc.complete <- hclust(distance.usarrests, method = "complete")
hc.average <- hclust(distance.usarrests, method = "average")
hc.single <- hclust(distance.usarrests, method = "single")
```

```{r hclust-usarrests-plots, fig.width=8, fig.height=7}
par(mfrow = c(3, 1), mar = c(3, 4, 3, 1))
plot(hc.complete, main = "Complete linkage", xlab = "", sub = "")
plot(hc.average, main = "Average linkage", xlab = "", sub = "")
plot(hc.single, main = "Single linkage", xlab = "", sub = "")
par(mfrow = c(1, 1))
```

The same observations can appear in visibly different positions in the three
dendrograms because the linkage rule changes which merge is considered closest.
A dendrogram is not a final partition until we choose a height or a number of
clusters at which to cut it.

## Cutting the dendrogram

```{r hclust-cuts}
complete_k4 <- cutree(hc.complete, k = 4)
average_k4 <- cutree(hc.average, k = 4)

head(complete_k4)
table(complete_k4)
table(average_k4)
```

```{r hclust-comparison}
comparison <- table(
  Complete_linkage = complete_k4,
  Average_linkage = average_k4
)
comparison
```

The cross-tabulation is a descriptive comparison, not a claim that one
clustering is correct. If two methods place the same states together, that
agreement suggests a stable grouping. Disagreement identifies observations
whose group membership depends on the linkage choice.

## A reproducible cluster interpretation

```{r hclust-profiles}
complete_profiles <- aggregate(
  USArrests,
  by = list(Cluster = complete_k4),
  FUN = mean
)
complete_profiles
```

Cluster profiles translate an abstract dendrogram into the original units. A
cluster with a high mean for `Murder` and `Assault`, for example, represents
states with higher average values on those measures in this dataset. It is
important to describe these as profiles rather than causal explanations.

# $K$-Means Clustering

## The objective

$K$-means requires us to choose the number of clusters $K$ in advance. It then
assigns each observation to one cluster and chooses cluster centres to minimise
within-cluster sum of squares:

$$
\operatorname{WSS}(C_1,\ldots,C_K)
= \sum_{k=1}^{K}\sum_{i\in C_k}
\|x_i-\bar x_k\|^2.
$$

The algorithm alternates between two steps:

1. assign each observation to its nearest current centre;
2. recompute each centre as the mean of the observations assigned to it.

Each step cannot increase the objective, but the procedure can stop at a local
minimum. Consequently, `nstart` runs the algorithm from multiple random
initialisations and keeps the solution with the smallest WSS.

## Applying $K$-means to `USArrests`

```{r kmeans-usarrests}
set.seed(1)
kmeans_3 <- kmeans(USArrests_scaled, centers = 3, nstart = 20)
kmeans_3$size
kmeans_3$centers
```

The centres are expressed in standardized units, so a centre of $1$ means
approximately one standard deviation above the overall feature mean. The
cluster sizes show whether the solution is balanced or dominated by one large
group.

```{r kmeans-usarrests-profiles}
kmeans_profiles <- aggregate(
  USArrests,
  by = list(Cluster = kmeans_3$cluster),
  FUN = mean
)
kmeans_profiles
```

Unlike hierarchical clustering, $K$-means does not produce all possible group
sizes in one fitted object. We must fit separate models for different values
of $K$ if we want to compare them.

## Choosing $K$ with an elbow plot

For each $K$, the total within-cluster sum of squares measures compactness. It
will always decrease as $K$ grows, because more centres give the algorithm
more flexibility. We therefore look for an elbow: a point after which adding
another cluster produces only modest improvement.

```{r kmeans-elbow}
set.seed(1)
K_values <- 1:8
wss_values <- sapply(K_values, function(K) {
  kmeans(USArrests_scaled, centers = K, nstart = 20)$tot.withinss
})

plot(K_values, wss_values, type = "b", pch = 19,
     xlab = "Number of clusters K",
     ylab = "Total within-cluster sum of squares",
     main = "USArrests: K-means elbow plot")
```

The elbow is a judgement aid rather than a mechanical proof. In an
unsupervised problem there is no response variable that can certify the best
choice, so subject-matter usefulness and stability should also influence the
choice.

# The `NCI60` Gene-Expression Data

## Data structure

The `NCI60` data contain gene-expression measurements for 64 cancer cell lines
across thousands of genes. The data object has a numeric matrix and a label
indicating the cancer type. We will use the labels for interpretation and
checking, but unsupervised methods are fitted using the expression matrix.

```{r nci60-load}
data(NCI60)
str(NCI60)
dim(NCI60$data)
table(NCI60$labs)
```

The raw number of genes is much larger than the number of cell lines. PCA is
therefore a natural first step: it gives a low-dimensional view of the
variation before clustering.

## PCA for `NCI60`

```{r nci60-pca}
nci60.pca <- prcomp(NCI60$data, scale. = TRUE)
summary(nci60.pca)
```

```{r nci60-pca-plot, fig.width=7, fig.height=5}
nci60_scores <- nci60.pca$x[, 1:2]
nci60_labels <- sort(unique(as.character(NCI60$labs)))
nci60_label_codes <- match(as.character(NCI60$labs), nci60_labels)
plot(nci60_scores,
  col = nci60_label_codes, pch = 19,
     xlab = "First principal component",
     ylab = "Second principal component",
     main = "NCI60 cell lines in the first two PCs")
legend("topright", legend = nci60_labels,
    col = seq_along(nci60_labels), pch = 19,
       cex = 0.65, bty = "n")
```

The colours are added after the unsupervised computation to see whether known
cancer labels line up with visible structure. Strong overlap would mean that
the first two PCs do not separate those types cleanly; visible grouping would
suggest that the expression measurements contain type-related structure.

## Hierarchical clustering for `NCI60`

```{r nci60-hclust}
nci60_distance <- dist(NCI60$data)
nci60_hc <- hclust(nci60_distance, method = "complete")

plot(nci60_hc, labels = NCI60$labs,
     main = "NCI60: complete-linkage clustering",
     xlab = "Cell lines", sub = "")
```

```{r nci60-hclust-compare}
nci60_clusters <- cutree(nci60_hc, k = 4)
table(Cluster = nci60_clusters, Cancer_type = NCI60$labs)
```

The table compares the unsupervised cluster assignments with the known labels.
It is not a training confusion matrix: the labels were not used to create the
clusters. Some clusters may contain several cancer types, and one cancer type
may appear in multiple clusters, because the expression structure need not
match the supplied labels exactly.

## $K$-means for `NCI60`

```{r nci60-kmeans}
set.seed(1)
nci60_kmeans <- kmeans(NCI60$data, centers = 4, nstart = 50)

table(Cluster = nci60_kmeans$cluster, Cancer_type = NCI60$labs)
```

The choice of $K=4$ is made here for comparison with the hierarchical cut. In
practice, we would examine several values of $K$, use stability or validation
criteria, and consider the scientific purpose of the grouping.

# Comparing the Methods

The methods can be summarised by the question each one answers:

| Method | Main question | Main tuning choice |
|:--|:--|:--|
| PCA | Which directions explain the most variation? | Number of components |
| Hierarchical clustering | How do observations merge at every distance level? | Distance, linkage, and cut |
| $K$-means | Which $K$ compact groups minimise within-cluster variation? | $K$ and initialisation |

A responsible analysis reports the choices that produced the result. Scaling,
distance, linkage, random seed, and the number of clusters are not cosmetic
details; they define the geometry of the unsupervised problem.

# Conclusion

PCA compresses correlated variables into orthogonal directions ordered by
explained variance. Hierarchical clustering provides a nested picture of
similarity, while $K$-means produces a direct partition for a selected $K$.
The `USArrests` example shows how scaling and linkage affect a clustering
result, and the `NCI60` example shows how PCA and clustering can reveal
structure in high-dimensional gene-expression data.

Because unsupervised learning has no response variable to score directly, the
best practice is to examine several reasonable specifications, report the
sensitivity of the conclusions, and interpret the resulting patterns using
subject-matter knowledge rather than treating one algorithmic output as an
objective truth.
