---
output:
  pdf_document:
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{amsmath}
  - \input{../../_common/preamble.tex}
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
```

```{=latex}
\makecoverpage{Chapter 4 --- Classification \\ Lab}
```

# Introduction

Linear regression predicts a *quantitative* response. But plenty of
interesting questions have a *qualitative* answer instead — will the
market go up or down today, will a customer buy the product or not, does
this email look like spam. This lab works through the main tools for that
kind of problem: logistic regression, linear and quadratic discriminant
analysis (LDA/QDA), naive Bayes, and $K$-nearest neighbors, applied to
three real data sets — daily S&P 500 returns, caravan insurance purchases,
and hourly bike-share demand.

```{r libraries}
library(ISLR2)
library(MASS)
library(class)
library(e1071)
```

# The Stock Market Data

`Smarket` records the percentage return of the S&P 500 on 1,250 trading
days (2001–2005), along with the previous five days' returns (`Lag1`
through `Lag5`), trading `Volume`, and `Direction` — whether the market
went `Up` or `Down` that day. The question we're chasing: can *yesterday's*
(and the last few days') returns predict *today's* direction at all?

```{r smarket-explore}
dim(Smarket)
summary(Smarket)
cor(Smarket[, -9])
```

The correlations between the lag variables and `Today`'s return all sit
close to zero — there's essentially no linear relationship between recent
past returns and today's. That's actually the expected, textbook result
for daily stock returns (a market that were linearly predictable from its
own recent past would be an unusually exploitable one). The only sizable
correlation in the table is between `Year` and `Volume` — unsurprising,
since trading volume trended upward over these five years:

```{r volume-plot, fig.width=6, fig.height=4}
plot(Smarket$Volume, xlab = "Chronological index", ylab = "Volume",
     main = "Trading volume increases over 2001-2005")
```

# Logistic Regression

Since `Direction` is binary, `lm()` isn't the right tool — nothing stops a
linear model from predicting a probability below 0 or above 1. Logistic
regression instead models the probability itself through the *logistic
function*, which is mathematically guaranteed to stay in $(0,1)$ no matter
what the predictors are:

$$
p(X) = P(\text{Direction = Up} \mid X) = \frac{e^{\beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p}}{1 + e^{\beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p}}
$$

Rearranging shows *why* it's called "logistic" — it's linear not in the
probability itself, but in the **log-odds**:

$$
\log\left(\frac{p(X)}{1 - p(X)}\right) = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p
$$

So a coefficient $\beta_j$ is read as: a one-unit increase in $X_j$
changes the log-odds of `Up` by $\beta_j$, holding everything else fixed —
not a direct change in probability, since the logistic curve is flattest
at the extremes and steepest near $p=0.5$.

```{r logistic-fit}
glm.fits <- glm(
  Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume,
  data = Smarket, family = binomial
)
summary(glm.fits)
```

None of the p-values are small — by itself, this is a first sign that
yesterday's returns don't carry much predictive signal for today's
direction, consistent with what the correlation matrix already suggested.
`Lag1` has the smallest p-value ($\approx 0.145$) with a *negative*
coefficient, which would (weakly) suggest that a positive return
yesterday makes a positive return today slightly less likely — but nowhere
near significant on its own.

## Confusion matrix and training accuracy

```{r logistic-confusion}
glm.probs <- predict(glm.fits, type = "response")
glm.pred <- rep("Down", nrow(Smarket))
glm.pred[glm.probs > 0.5] <- "Up"

table(glm.pred, Smarket$Direction)
mean(glm.pred == Smarket$Direction)
```

About 52% correct — barely better than a coin flip, and misleadingly so:
this is **training** accuracy, computed on the exact data the model was
fit to. A model that just memorizes noise in the training set can look
fine here and still fail on new data — which is exactly why we need a
genuine held-out test set.

## A real train/test split

We use the classic "hold out the future" split for time series data:
train on 2001–2004, and test on 2005 — data the model has never seen.

```{r logistic-holdout}
train <- Smarket$Year < 2005
Smarket.2005 <- Smarket[!train, ]
Direction.2005 <- Smarket$Direction[!train]
dim(Smarket.2005)

glm.fits <- glm(
  Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume,
  data = Smarket, family = binomial, subset = train
)
glm.probs <- predict(glm.fits, Smarket.2005, type = "response")
glm.pred <- rep("Down", nrow(Smarket.2005))
glm.pred[glm.probs > 0.5] <- "Up"

table(glm.pred, Direction.2005)
mean(glm.pred == Direction.2005)
```

Test accuracy is *worse* than training accuracy here — actually worse
than just always guessing `Up`, which is a sign the full five-lag model is
overfitting noise rather than finding real signal. Since `Lag1` and
`Lag2` were the two most promising predictors above, it's worth trying a
leaner model:

```{r logistic-lean}
glm.fits <- glm(Direction ~ Lag1 + Lag2, data = Smarket,
                 family = binomial, subset = train)
glm.probs <- predict(glm.fits, Smarket.2005, type = "response")
glm.pred <- rep("Down", nrow(Smarket.2005))
glm.pred[glm.probs > 0.5] <- "Up"

table(glm.pred, Direction.2005)
mean(glm.pred == Direction.2005)
```

That's a meaningful improvement — dropping the three weakest predictors
actually helped test performance, a concrete illustration of the
bias-variance idea from Chapter 2: a model with fewer, better-chosen
predictors can generalize better than one that includes everything,
even though it necessarily has strictly worse (or equal) training fit.

# Linear Discriminant Analysis

Logistic regression models $P(Y \mid X)$ directly. LDA takes the opposite
route — Bayes' theorem in reverse — modeling how $X$ is distributed
*within* each class, then using Bayes' rule to get back to $P(Y \mid X)$:

$$
P(Y = k \mid X = x) = \frac{\pi_k \, f_k(x)}{\sum_{l} \pi_l \, f_l(x)}
$$

where $\pi_k$ is the (estimated) prior probability of class $k$, and
$f_k(x)$ is the density of $X$ within class $k$. LDA's specific assumption
is that each class's predictors follow a Gaussian distribution, all
sharing the *same* covariance structure — only the class means differ.
Under that assumption the decision boundary between classes works out to
be linear, which is where LDA gets its name.

```{r lda-fit}
lda.fit <- lda(Direction ~ Lag1 + Lag2, data = Smarket, subset = train)
lda.fit
```

The **prior probabilities** ($\hat\pi_{\text{Down}} \approx 0.492$,
$\hat\pi_{\text{Up}} \approx 0.508$) are just the class proportions in the
training data. The **group means** show that on `Up` days, `Lag1` and
`Lag2` tend to be slightly negative on average — consistent with the
(weak) negative association logistic regression also picked up.

```{r lda-predict}
lda.pred <- predict(lda.fit, Smarket.2005)
table(lda.pred$class, Direction.2005)
mean(lda.pred$class == Direction.2005)
```

Almost identical accuracy to the leaner logistic regression — not a
coincidence. With two classes and (roughly) Gaussian, similarly-shaped
predictors, LDA and logistic regression tend to land on very similar
decision boundaries, even though they get there through different routes
(generative vs. discriminative modeling).

# Quadratic Discriminant Analysis

QDA relaxes LDA's shared-covariance assumption — each class gets its
*own* covariance matrix. That extra flexibility means the decision
boundary is now quadratic rather than linear, at the cost of estimating
more parameters (and so needing more data to do it reliably).

```{r qda-fit}
qda.fit <- qda(Direction ~ Lag1 + Lag2, data = Smarket, subset = train)
qda.fit

qda.class <- predict(qda.fit, Smarket.2005)$class
table(qda.class, Direction.2005)
mean(qda.class == Direction.2005)
```

QDA edges out both logistic regression and LDA here. One reading: the
true boundary between `Up` and `Down` days may genuinely be closer to
quadratic than linear — but with only two predictors and a few hundred
test observations, this could also just be QDA getting fortunate on this
particular test split rather than a guaranteed generalizable edge.

# Naive Bayes

Naive Bayes keeps the same generative Bayes'-rule structure as LDA/QDA,
but replaces the Gaussian-with-shared/class-specific-covariance assumption
with something blunter and computationally cheaper: it assumes every
predictor is **independent of every other predictor, within each class**.

$$
f_k(x) = f_{k1}(x_1) \times f_{k2}(x_2) \times \cdots \times f_{kp}(x_p)
$$

This is rarely *exactly* true — `Lag1` and `Lag2` are unlikely to be truly
independent given `Direction` — but it collapses a hard multivariate
density-estimation problem into $p$ easy univariate ones, and in practice
often still classifies well even when the independence assumption is
technically wrong.

```{r naive-bayes}
nb.fit <- naiveBayes(Direction ~ Lag1 + Lag2, data = Smarket, subset = train)
nb.fit

nb.class <- predict(nb.fit, Smarket.2005)
table(nb.class, Direction.2005)
mean(nb.class == Direction.2005)
```

Performance here lands close to LDA and logistic regression — reasonable,
since with only two predictors there's less room for the independence
assumption to cause real damage than there would be with many correlated
predictors.

# $K$-Nearest Neighbors

Every method so far assumes some parametric form for how `Direction`
depends on the predictors (a linear log-odds, a Gaussian density, ...).
KNN assumes nothing about the shape of the boundary at all: to classify a
new point, find its $K$ nearest neighbors in the training data (by
Euclidean distance) and take a majority vote among them.

```{r knn-smarket}
train.X <- cbind(Smarket$Lag1, Smarket$Lag2)[train, ]
test.X <- cbind(Smarket$Lag1, Smarket$Lag2)[!train, ]
train.Direction <- Smarket$Direction[train]

set.seed(1)
knn.pred <- knn(train.X, test.X, train.Direction, k = 1)
table(knn.pred, Direction.2005)
mean(knn.pred == Direction.2005)
```

$K=1$ is the most flexible possible choice — each test point is classified
by whichever single training point happens to be closest, which chases
noise aggressively and shows in the roughly 50% accuracy (no better than
guessing). Bumping $K$ up trades away some of that flexibility for
stability:

```{r knn-k3}
knn.pred <- knn(train.X, test.X, train.Direction, k = 3)
table(knn.pred, Direction.2005)
mean(knn.pred == Direction.2005)
```

A modest improvement, but still well short of QDA. Altogether, across
every method tried, `Lag1`/`Lag2` simply don't carry much signal about
`Direction` — which, again, is the expected finding for daily stock
returns, not a failure of the methods themselves.

## Application: Caravan Insurance

`Caravan` is a more favorable setting for KNN: 5,822 customers, 85
demographic predictors, and a binary `Purchase` outcome (did they buy
caravan insurance). Because KNN measures raw Euclidean distance, any
predictor on a larger numeric scale automatically dominates the distance
calculation regardless of how informative it actually is — so predictors
need to be standardized first:

```{r caravan-scale}
standardized.X <- scale(Caravan[, -86])
mean(Caravan[[1]]); mean(Caravan[[2]])
var(standardized.X[, 1]); var(standardized.X[, 2])
```

After scaling, every predictor has variance 1, so no single variable can
dominate the distance metric purely because of its units.

```{r caravan-knn}
test <- 1:1000
train.X <- standardized.X[-test, ]
test.X <- standardized.X[test, ]
train.Y <- Caravan$Purchase[-test]
test.Y <- Caravan$Purchase[test]

set.seed(1)
knn.pred <- knn(train.X, test.X, train.Y, k = 3)
table(knn.pred, test.Y)
```

Only about 6% of customers actually purchase, so overall accuracy is a
misleading metric here — a model that always predicts "No" would already
be right about 94% of the time. What actually matters for a marketing
campaign is *precision among predicted buyers*: of the customers KNN
flagged as likely purchasers, what fraction really did buy? That's a
meaningfully higher hit rate than the 6% baseline rate, which is the
actual value of the model — it concentrates marketing spend on
customers substantially more likely to convert than a random customer
would be.

# Poisson Regression

`Bikeshare` records hourly bike rental counts (`bikers`) along with
weather and calendar predictors. A count that's always $\geq 0$ and often
small doesn't fit the assumptions of ordinary linear regression well — a
linear model can happily predict a negative number of bike rentals, which
is nonsensical. Poisson regression is built for exactly this: it assumes
$Y \mid X \sim \text{Poisson}(\lambda(X))$, and — just like logistic
regression links probability through the logistic function — links the
rate $\lambda$ through a **log** link, guaranteeing predictions stay
positive:

$$
\log(\lambda(X)) = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p
\qquad \Longleftrightarrow \qquad
\lambda(X) = e^{\beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p}
$$

A coefficient $\beta_j$ is therefore read multiplicatively: a one-unit
increase in $X_j$ multiplies the expected count by $e^{\beta_j}$, holding
other predictors fixed.

```{r poisson-fit}
mod.pois <- glm(
  bikers ~ mnth + hr + workingday + temp + weathersit,
  data = Bikeshare, family = poisson
)
summary(mod.pois)
```

```{r poisson-hourly, fig.width=7, fig.height=4}
coef.hr <- c(0, coef(mod.pois)[grepl("^hr", names(coef(mod.pois)))])
plot(0:23, coef.hr, type = "l", pch = 19, col = "steelblue", lwd = 2,
     xlab = "Hour of the day", ylab = "Coefficient (log rental rate)",
     main = "Bikeshare demand by hour, holding other predictors fixed")
```

The hourly coefficients trace out the expected daily commuting rhythm —
sharp peaks around the morning and evening rush hours, with a trough
overnight — which a positive-count model like this one can represent
naturally, in a way a plain linear regression fit to the raw counts
would not.
