---
output:
  pdf_document:
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{amsmath}
  - \input{../_common/preamble.tex}
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
```

```{=latex}
\makecoverpage{Chapter 2 --- Statistical Learning \\ Exercises 1 and 10}
```

# Exercise 1 — Flexible vs. Inflexible Methods

For every part below, the comparison comes down to one identity: the expected
test mean squared error (MSE) at a new point $x_0$ decomposes as

$$
E\Big[(y_0 - \hat{f}(x_0))^2\Big] = \text{Var}\big(\hat{f}(x_0)\big) + \big[\text{Bias}\big(\hat{f}(x_0)\big)\big]^2 + \text{Var}(\epsilon)
$$

A **flexible** method (e.g. splines, high-degree polynomials, kNN with small
$k$) can bend to follow intricate patterns in the training data, so it tends
to have **low bias** but **high variance** — it changes a lot if you swap in
a different training sample. An **inflexible** method (e.g. linear
regression) is the opposite: it can't bend much, so it has **higher bias**
but **low variance** — it barely changes from one training sample to the
next. Every part of this question is really asking: *in this scenario, does
reducing bias (going flexible) buy you more than the variance it costs?*

## (a) $n$ extremely large, $p$ small

**Better: flexible.**

Think of it like sketching a coastline. With only a handful of GPS points (small
$n$), you can only draw a straight or gently curved line — try to trace every
inlet and you'll just be connecting noise. But hand a cartographer thousands
of closely-spaced points (large $n$), and tracing every inlet and headland
becomes not just possible but *more* accurate than forcing a straight line
through them.

Formally: a large $n$ relative to $p$ means there is enough data to estimate
a complex $\hat{f}$ **without** the fit becoming unstable — the
$\text{Var}(\hat{f}(x_0))$ term stays under control even as flexibility goes
up, because each local region of the predictor space is still densely
populated with observations. Meanwhile an inflexible method leaves the bias
term unnecessarily large when there's no need for it — there's enough data
to support a far better approximation of the true $f$. So the bias reduction
from flexibility comes almost for free.

## (b) $p$ extremely large, $n$ small

**Better: inflexible.**

This is the mirror image of (a), and it's the classic *curse of
dimensionality*: imagine trying to furnish a house (small $n$ = few pieces of
furniture) that keeps gaining extra rooms (large $p$ = more predictors).
Spread thin across so many rooms, the house starts to look empty no matter
how you arrange things — there just isn't enough "stuff" to fill the space
convincingly.

Formally: with few observations spread across a high-dimensional predictor
space, each region of that space is sparsely populated. A flexible method
will latch onto whatever idiosyncratic pattern happens to exist among those
few points — $\text{Var}(\hat{f}(x_0))$ explodes — and that variance
inflation outweighs any bias it saves. An inflexible method, constrained to a
simple global shape, can't overfit the sparse data in the same way, so it
generalizes better despite its higher bias.

## (c) Relationship between predictors and response is highly non-linear

**Better: flexible.**

Picture trying to trace a winding mountain road using only straight
segments of highway — you'll cut every corner, and the further the road
curves, the worse your straight-line approximation gets. A flexible method
is the one willing to bend with the road.

Formally: if the true $f$ is highly non-linear, an inflexible method like
linear regression has a *structural* limitation — no matter how much data
you give it, it cannot represent the curvature, so
$\big[\text{Bias}\big(\hat{f}(x_0)\big)\big]^2$ stays large regardless of
$n$. A flexible method can actually approximate the curvature, driving bias
down. As long as $n$ is reasonably sized, the resulting increase in variance
is smaller than the bias it eliminates, so total expected test MSE drops.

## (d) Variance of the error terms $\sigma^2 = \text{Var}(\epsilon)$ is extremely high

**Better: inflexible.**

Think of trying to hear a friend's voice (the true signal $f$) over a
room full of loud static (the irreducible noise $\epsilon$). Leaning in and
straining to catch every fluctuation in the sound (a flexible method) mostly
means you start "hearing" patterns in the static itself, not the voice. You're
better off just catching the general gist.

Formally: $\text{Var}(\epsilon)$ is **irreducible** — it sits in the MSE
decomposition as a constant floor that *no* method, however flexible, can
lower. A flexible method doesn't just fail to reduce it — it actively makes
things worse, because it will interpret large random fluctuations in the
training data as if they were real signal and fit to them, inflating
$\text{Var}(\hat{f}(x_0))$ on top of the noise floor that was already there.
An inflexible method resists chasing that noise, so its total expected test
MSE stays closer to the irreducible floor.

\newpage

# Exercise 10 — The `Boston` Housing Data Set

```{r load-data}
library(ISLR2)
data(Boston)
```

## (a) Rows, columns, and what they represent

```{r dim-boston}
dim(Boston)
```

The data set has **506 rows and 13 columns**. Each **row** is a single
census tract (a small geographic subdivision) in the Boston area. Each
**column** is a variable recorded for that tract — for example `crim` (per
capita crime rate), `rm` (average rooms per dwelling), `ptratio`
(pupil-teacher ratio), and `medv` (median value of owner-occupied homes in
\$1000s), among others. So this is not 506 individual houses — it's 506
neighborhoods, each summarized by 13 socioeconomic and structural
statistics.

## (b) Pairwise scatterplots

```{r pairs-plot, fig.width=9, fig.height=9}
pairs(Boston, pch = 20, cex = 0.4)
```

A full $13\times13$ grid is dense, so it helps to know what to look for
rather than eyeball all 78 panels equally. A few patterns stand out
visually and are confirmed by the correlations pulled out in part (c)
below:

- **`lstat` vs. `medv`**: a clear negative, curved (not perfectly linear)
  relationship — tracts with a higher share of lower-status population
  have systematically lower home values.
- **`rm` vs. `medv`**: a positive relationship — more rooms per dwelling
  tends to mean higher home value, which makes intuitive sense (bigger
  houses cost more).
- **`nox` vs. `indus`** and **`nox` vs. `age`**: positive relationships —
  tracts with more industrial land use, and with older housing stock, tend
  to have higher nitrogen oxide pollution.
- **`tax` vs. `rad`**: a striking pattern where many tracts cluster at one
  very high value of `rad` (the accessibility-to-highways index) paired
  with a high, fairly constant `tax` rate — visible as a dense vertical/
  horizontal band rather than a smooth trend.
- **`dis` vs. `nox`, `age`, `indus`**: negative relationships — tracts
  farther from Boston's employment centers tend to be less industrial,
  have newer housing, and lower pollution (consistent with them being
  more suburban).

## (c) Predictors associated with per-capita crime rate

```{r crim-corr}
sort(cor(Boston)[, "crim"])
```

Yes — several predictors show a meaningful relationship with `crim`:

- **`rad`** (r $\approx$ 0.63) and **`tax`** (r $\approx$ 0.58) are the
  strongest — tracts with better highway accessibility and higher property
  tax rates tend to have notably higher crime rates. This is really one
  pattern more than two, since `rad` and `tax` are themselves highly
  correlated (visible as that dense cluster in part (b)) — a small subset
  of tracts with maxed-out `rad` and `tax` values are dragging crime up
  with them.
- **`lstat`** (r $\approx$ 0.46) and **`nox`** (r $\approx$ 0.42) are
  moderately positively associated — tracts with more lower-status
  population and more air pollution tend to have higher crime.
- **`medv`** (r $\approx$ -0.39) and **`dis`** (r $\approx$ -0.38) are
  moderately negatively associated — higher home values and greater
  distance from employment centers both correspond to lower crime.

Note these are all linear (Pearson) correlations; the true relationships
may well be non-linear, so these numbers indicate *association*, not
necessarily the full shape of the relationship.

## (d) Notably high crime, tax, or pupil-teacher tracts; ranges

```{r ranges}
sapply(Boston[, c("crim", "tax", "ptratio")], range)
sapply(Boston[, c("crim", "tax", "ptratio")], summary)
```

- **`crim`** ranges from **0.006 to 88.98** — an enormous spread. The
  median tract has a crime rate near 0.26, so a handful of tracts (crime
  rate in the 40s–80s) are extreme outliers pulling the mean well above
  the median — most tracts are low-crime, with a small number of very
  high-crime tracts.
- **`tax`** ranges from **187 to 711** (per \$10,000 of property value).
  There's a visible cluster of tracts sitting right at the maximum (666),
  the same cluster seen in the `tax`–`rad` pattern from part (b) — so while
  the *range* is wide, a meaningful subgroup of tracts is bunched at the
  top of it.
- **`ptratio`** ranges from **12.6 to 22.0** — a comparatively narrow
  spread on an absolute scale, though a pupil-teacher ratio of 22 vs. 12.6
  is still a substantively large difference in school resourcing per
  student.

## (e) Tracts bounding the Charles River

```{r chas-sum}
sum(Boston$chas)
```

**35** of the 506 census tracts bound the Charles River (`chas == 1`).

## (f) Median pupil-teacher ratio

```{r ptratio-median}
median(Boston$ptratio)
```

The median pupil-teacher ratio across tracts is **19.05**.

## (g) Tract with the lowest median home value

```{r min-medv}
which.min(Boston$medv)
Boston[which.min(Boston$medv), ]
```

Tract **399** has the lowest `medv` at **\$5,000** — far below the data
set's minimum-to-typical range (the next visualization makes the point
concretely: this tract sits at the very bottom of the `medv` distribution
while most other predictors sit near the *top* of their own ranges).

```{r min-medv-percentile}
idx <- which.min(Boston$medv)
sapply(names(Boston), function(v) round(100 * mean(Boston[[v]] <= Boston[idx, v]), 1))
```

Interpreting these as percentile ranks within the full data set, tract 399
is a striking outlier profile: it sits at the **100th percentile for
`age`** (essentially all owner-occupied units built before 1940) and
**`rad`** (maximum highway accessibility), the **99th percentile for
`tax`**, the **98.8th percentile for `crim`**, and the **97.8th percentile
for `lstat`** — while its `rm` (rooms per dwelling) is only at the 7.7th
percentile and its `dis` (distance to employment centers) at the 5.7th
percentile. In short: old housing stock, very high crime and tax burden,
small homes, low-status population, and close to the urban core — a
textbook combination for depressed home values, and consistent with the
associations found in part (c).

## (h) Tracts averaging more than seven / eight rooms

```{r rooms}
sum(Boston$rm > 7)
sum(Boston$rm > 8)
summary(Boston[Boston$rm > 8, ])
```

**64** tracts average more than seven rooms per dwelling, and **13** average
more than eight.

The 13 tracts with `rm > 8` are a notably *advantaged* subset compared to
the data set as a whole: their median `crim` ($\approx$ 0.52) and `lstat`
($\approx$ 4.14) are both far below the full data set's medians, and their
median `medv` (\$48,300) sits near the top of the overall home-value range
(recall the overall `medv` maximum is \$50,000, likely censored at that
value). In other words, large-home tracts tend to combine low crime, low
lower-status population share, and high home values — consistent with the
negative `lstat`–`medv` and positive `rm`–`medv` relationships seen in part
(b).
