---
output:
  pdf_document:
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{amsmath}
  - \input{../../_common/preamble.tex}
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
```

```{=latex}
\makecoverpage{Chapter 5 --- Resampling Methods \\ Exercises 5 and 9}
```

```{r libraries}
library(ISLR2)
library(boot)
```

# Exercise 5 — Validation-Set Error for Logistic Regression on `Default`

## (a) Fit the full-data logistic regression

```{r ex5-full-fit}
set.seed(1)
glm.fit <- glm(default ~ income + balance, data = Default, family = binomial)
summary(glm.fit)
```

Both `income` and `balance` come out highly significant, with `balance`
the far stronger predictor by t-statistic magnitude — consistent with what
Chapter 4 found for this same relationship.

## (b) Validation-set test error

The validation-set approach is conceptually simple: split the data in
half, fit only on one half, and check performance on the half the model
never saw — the resulting error rate is a genuine estimate of test error,
unlike the training error computed on data the model already fit.

```{r ex5-validation}
set.seed(1)
n <- nrow(Default)
train <- sample(n, n / 2)

glm.fit <- glm(default ~ income + balance, data = Default,
                family = binomial, subset = train)
glm.probs <- predict(glm.fit, Default[-train, ], type = "response")
glm.pred <- ifelse(glm.probs > 0.5, "Yes", "No")

val.error <- mean(glm.pred != Default[-train, ]$default)
val.error
```

A validation error around 2.5% — notably higher than the essentially-zero
training error a full-data fit could give the illusion of, since 5,000
held-out observations never influenced the fitted coefficients at all.

## (c) Repeating with three different splits

```{r ex5-repeat}
val_error <- function(seed) {
  set.seed(seed)
  train <- sample(n, n / 2)
  fit <- glm(default ~ income + balance, data = Default,
             family = binomial, subset = train)
  probs <- predict(fit, Default[-train, ], type = "response")
  pred <- ifelse(probs > 0.5, "Yes", "No")
  mean(pred != Default[-train, ]$default)
}

errors <- sapply(c(1, 2, 3), val_error)
errors
```

The three error rates all land in a fairly tight band (roughly 2.4%–2.6%),
but they're not identical — which is the point. Every one of these
numbers is a valid estimate of the same underlying test error, yet each
depends on the arbitrary luck of which 5,000 observations happened to
land in the training half. That split-to-split variability is exactly the
validation set approach's core weakness: throwing away half the data for
training, on a single random split, means both the fitted model and the
resulting error estimate are noisier than they'd need to be with a method
like $k$-fold cross-validation that reuses the data more efficiently.

## (d) Adding a `student` dummy variable

```{r ex5-student}
set.seed(1)
train <- sample(n, n / 2)
glm.fit.student <- glm(default ~ income + balance + student, data = Default,
                        family = binomial, subset = train)
glm.probs <- predict(glm.fit.student, Default[-train, ], type = "response")
glm.pred <- ifelse(glm.probs > 0.5, "Yes", "No")

val.error.student <- mean(glm.pred != Default[-train, ]$default)
val.error.student
```

With the same seed (and so the same train/test split) as (b), adding
`student` gives a validation error of `r round(val.error.student, 4)`,
essentially identical to — in this run, marginally *higher* than —
`r round(val.error, 4)` without it. So there's no evidence here that
including `student` improves predictive performance. This isn't
surprising in hindsight: `student` is strongly correlated with `balance`
(students tend to carry different balances than non-students), so much of
whatever information `student` carries about default risk is already
captured once `balance` is in the model — adding it doesn't give logistic
regression much new signal to work with, only extra estimation noise.

\newpage

# Exercise 9 — Bootstrap Estimates for the `Boston` Data

Every part below is about the same core idea: `medv` (or a statistic of
it) is being estimated from one particular sample of 506 census tracts,
and we want to know how much that estimate would bounce around if we
happened to have drawn a different sample of the same size from the same
population. Textbook formulas exist for some statistics (the sample
mean); the bootstrap gives us an answer even for the ones that don't (the
median, a percentile) by treating our one sample as a stand-in for the
population and resampling from it directly.

## (a) Estimate of the population mean, $\hat\mu$

```{r ex9-mu}
Boston2 <- ISLR2::Boston
mu.hat <- mean(Boston2$medv)
mu.hat
```

## (b) Standard error of $\hat\mu$, by formula

$$
\text{SE}(\hat\mu) = \frac{s}{\sqrt{n}}
$$

where $s$ is the sample standard deviation of `medv` and $n = 506$.

```{r ex9-se-formula}
se.formula <- sd(Boston2$medv) / sqrt(nrow(Boston2))
se.formula
```

This measures how much $\hat\mu$ would be expected to vary, on average,
across repeated samples of 506 tracts drawn from the same population —
about \$409 (recall `medv` is in \$1000s), against a mean around \$22,530.

## (c) Standard error of $\hat\mu$, by bootstrap

```{r ex9-boot-mean}
set.seed(1)
boot.fn.mean <- function(data, index) mean(data[index])
boot.mean <- boot(Boston2$medv, boot.fn.mean, R = 10000)
boot.mean
```

The bootstrap standard error (`r round(sd(boot.mean$t), 4)`) comes out
essentially identical to the formula-based estimate in (b)
(`r round(se.formula, 4)`) — reassuring, since for something as
well-behaved as a sample mean, the classical formula is already exact
asymptotically. The bootstrap isn't doing anything the formula couldn't
already do here; its real value shows up in (f) and (h), for statistics
that have no comparably simple formula at all.

## (d) 95% confidence interval for the mean, via bootstrap

Using the approximation given in the exercise, $[\hat\mu - 2\,\text{SE}(\hat\mu),\ \hat\mu + 2\,\text{SE}(\hat\mu)]$:

```{r ex9-ci}
se.boot <- sd(boot.mean$t)
ci.boot <- c(mu.hat - 2 * se.boot, mu.hat + 2 * se.boot)
ci.boot

t.test(Boston2$medv)
```

The bootstrap-based interval (roughly `r round(ci.boot[1],2)` to
`r round(ci.boot[2],2)`) is nearly indistinguishable from `t.test()`'s
interval — expected, since both are really estimating the same quantity
($\pm 2$ standard errors around the mean is a close approximation to the
$t$-distribution multiplier `t.test()` uses at $n=506$), just arrived at
through different machinery: one assumes normal-theory sampling
distribution math, the other estimates the same spread empirically by
actually resampling.

## (e) Estimate of the population median

```{r ex9-median}
med.hat <- median(Boston2$medv)
med.hat
```

## (f) Standard error of the median, via bootstrap

There is no simple closed-form formula for the standard error of a
sample median (unlike the mean, in (b)) — this is precisely the situation
the bootstrap is built for.

```{r ex9-boot-median}
set.seed(1)
boot.fn.median <- function(data, index) median(data[index])
boot.median <- boot(Boston2$medv, boot.fn.median, R = 10000)
boot.median
```

The estimated standard error (`r round(sd(boot.median$t), 4)`) is
somewhat *smaller* than the mean's standard error from (c) — the median
is less sensitive than the mean to the handful of very high-`medv`
tracts (recall `medv` appears capped at \$50,000 in this data), so it
varies a bit less across resamples.

## (g) Estimate of the 10th percentile

```{r ex9-q10}
mu.hat.01 <- quantile(Boston2$medv, 0.1)
mu.hat.01
```

## (h) Standard error of the 10th percentile, via bootstrap

```{r ex9-boot-q10}
set.seed(1)
boot.fn.q10 <- function(data, index) quantile(data[index], 0.1)
boot.q10 <- boot(Boston2$medv, boot.fn.q10, R = 10000)
boot.q10
```

The standard error here (`r round(sd(boot.q10$t), 4)`) is the largest of
the three statistics examined — larger than both the mean's and the
median's. This tracks with intuition: the 10th percentile is estimated
from comparatively few, more extreme observations in the lower tail of
`medv`, so there's inherently less data locally supporting that estimate
than there is supporting the mean or median, which draw on the full
distribution. This is exactly the kind of finding the bootstrap makes
practical to obtain — no standard formula for the standard error of a
percentile exists at all, yet the same resampling procedure that worked
for the mean and median hands us an answer here just as directly.
