---
output:
  pdf_document:
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{amsmath}
  - \input{../../_common/preamble.tex}
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

```{=latex}
\makecoverpage{Chapter 5 --- Resampling Methods \\ Lab}
```

# Introduction

When a model is fitted, its training error is usually too optimistic: the
same observations were used both to estimate the model and to measure how
well it fits. The real questions are different. How well will the model
predict new observations? How much would its coefficients change if we had
collected a different sample? Which model complexity gives the best balance
between underfitting and overfitting?

Resampling methods answer these questions by repeatedly reusing the observed
data in carefully chosen ways. This lab develops four closely related ideas:

1. the validation-set approach;
2. leave-one-out cross-validation (LOOCV);
3. $k$-fold cross-validation; and
4. the bootstrap.

The first three estimate prediction error. The bootstrap estimates the
sampling variability of statistics such as regression coefficients and
portfolio weights. We use the `Auto` and `Portfolio` data sets from `ISLR2`.
The main prediction problem is to model automobile fuel efficiency (`mpg`)
from horsepower using polynomial regression.

```{r libraries}
library(ISLR2)
library(boot)
```

## The prediction problem and the bias--variance trade-off

Let $Y$ be a response and $X$ a predictor. A fitted model produces
$\hat{f}(X)$, and its prediction error on a new observation can be written as

$$
\mathbb{E}\left[(Y - \hat{f}(X))^2 \mid X=x\right]
= \underbrace{\sigma^2}_{\text{irreducible noise}}
+ \underbrace{\operatorname{Bias}\left[\hat{f}(x)\right]^2}_{\text{systematic error}}
+ \underbrace{\operatorname{Var}\left[\hat{f}(x)\right]}_{\text{sensitivity to the sample}}.
$$

A very simple model may have high bias because it cannot represent the true
relationship. A highly flexible model may have low training error but high
variance because it follows random details of the particular sample. The
purpose of resampling is to estimate the out-of-sample part of this trade-off
using only the data we have.

# Validation Set Approach

## Idea and mathematical definition

The validation-set approach randomly divides the data into a training set and
a validation set. We fit the model using only the training observations, then
evaluate predictions on observations that were not used for fitting. For
regression, the estimate is the validation mean squared error (MSE):

$$
\operatorname{MSE}_{\mathrm{val}}
= \frac{1}{n_{\mathrm{val}}}
\sum_{i \in \mathcal{V}}
\left(y_i - \hat{f}_{\mathcal{T}}(x_i)\right)^2,
$$

where $\mathcal{T}$ is the training set and $\mathcal{V}$ is the validation
set. The notation on $\hat{f}_{\mathcal{T}}$ matters: the model was fitted
using the training observations only.

The approach is easy and fast, but the result depends on one random split.
Also, because the model is trained on fewer observations than are ultimately
available, its validation error can be a little pessimistic.

## One training/validation split

`Auto` contains 392 cars after missing observations have been removed. We
compare polynomial models of degrees one, two, and three. A fixed seed makes
the random split reproducible for the report.

```{r validation-setup}
data(Auto)
Auto <- na.omit(Auto)
n <- nrow(Auto)

set.seed(1)
train <- sample(seq_len(n), size = n / 2)
validation <- -train

validation_mse <- function(model, data, validation_indices) {
  mean((data$mpg - predict(model, newdata = data))[validation_indices]^2)
}

validation_results <- data.frame(
  Degree = 1:3,
  Validation_MSE = c(
    validation_mse(lm(mpg ~ horsepower, data = Auto, subset = train), Auto, validation),
    validation_mse(lm(mpg ~ poly(horsepower, 2), data = Auto, subset = train), Auto, validation),
    validation_mse(lm(mpg ~ poly(horsepower, 3), data = Auto, subset = train), Auto, validation)
  )
)
validation_results
```

The model with the smallest validation MSE is preferred for this split. The
quadratic model normally improves substantially over the linear model because
fuel efficiency falls quickly at first and then levels off. If the cubic
model is not better, that is evidence that its extra flexibility is fitting
sample-specific variation rather than a reliable pattern.

```{r validation-plot, fig.width=6, fig.height=4}
plot(
  validation_results$Degree,
  validation_results$Validation_MSE,
  type = "b", pch = 19, col = "steelblue",
  xlab = "Polynomial degree", ylab = "Validation MSE",
  main = "Validation error for polynomial models"
)
```

## Why one split is not enough

The validation set is itself a random sample. Changing the split changes both
the fitted coefficients and the observations used to judge them. The model
comparison is therefore noisy even when the underlying data-generating
process has not changed.

```{r validation-repeated}
validation_mse_for_seed <- function(seed, degree) {
  set.seed(seed)
  training_indices <- sample(seq_len(n), size = n / 2)
  model_formula <- if (degree == 1) {
    mpg ~ horsepower
  } else {
    mpg ~ poly(horsepower, degree)
  }
  model <- lm(model_formula, data = Auto, subset = training_indices)
  mean(
    (Auto$mpg - predict(model, newdata = Auto))[-training_indices]^2
  )
}

split_results <- expand.grid(Seed = 1:10, Degree = 1:3)
split_results$Validation_MSE <- mapply(
  validation_mse_for_seed,
  split_results$Seed,
  split_results$Degree
)

aggregate(Validation_MSE ~ Degree, data = split_results,
          FUN = function(x) c(Mean = mean(x), SD = sd(x)))
```

The standard deviation across splits shows the weakness of relying on one
holdout set. Repeated splitting is informative, but cross-validation gives a
more systematic way to ensure that every observation participates in both
training and validation.

# Leave-One-Out Cross-Validation

## Procedure

LOOCV creates $n$ training sets. For the $i$th fit, observation $i$ is left
out, the model is fitted to the other $n-1$ observations, and the held-out
prediction error is recorded. The final estimate is

$$
\operatorname{CV}_{(n)}
= \frac{1}{n}\sum_{i=1}^{n}
\left(y_i - \hat{f}^{(-i)}(x_i)\right)^2,
$$

where $\hat{f}^{(-i)}$ means that observation $i$ was excluded when fitting.
Unlike a single validation split, every observation is used for validation
exactly once and for training $n-1$ times.

For linear regression there is a useful shortcut. If $e_i=y_i-\hat{y}_i$ is
the ordinary residual and $h_{ii}$ is the leverage of observation $i$, then

$$
\operatorname{CV}_{(n)}
= \frac{1}{n}\sum_{i=1}^{n}
\left(\frac{e_i}{1-h_{ii}}\right)^2.
$$

The `cv.glm()` function uses this idea for a generalized linear model. An
ordinary Gaussian `glm()` is equivalent to `lm()` for this purpose.

```{r loocv}
loocv_results <- data.frame(
  Degree = 1:10,
  LOOCV_MSE = NA_real_
)

for (degree in loocv_results$Degree) {
  model_formula <- if (degree == 1) {
    mpg ~ horsepower
  } else {
    mpg ~ poly(horsepower, degree)
  }
  model <- glm(model_formula, data = Auto)
  loocv_results$LOOCV_MSE[degree] <- cv.glm(Auto, model)$delta[1]
}

loocv_results
```

```{r loocv-plot, fig.width=6, fig.height=4}
plot(
  loocv_results$Degree, loocv_results$LOOCV_MSE,
  type = "b", pch = 19, col = "firebrick",
  xlab = "Polynomial degree", ylab = "LOOCV MSE",
  main = "Leave-one-out cross-validation"
)
```

The typical pattern is a large improvement from degree one to degree two,
followed by a flat curve. That means the quadratic term captures important
curvature, while higher-order terms do not produce a dependable reduction in
prediction error. LOOCV is attractive because each fit uses almost the full
data set, but it requires many model fits and can be computationally costly
for large data sets.

# $k$-Fold Cross-Validation

## Procedure and trade-off

In $k$-fold cross-validation, the observations are divided into $k$ folds.
Each fold is held out once while the other $k-1$ folds are used for training:

$$
\operatorname{CV}_{(k)}
= \frac{1}{k}\sum_{j=1}^{k}
\left(\frac{1}{|\mathcal{V}_j|}
\sum_{i\in\mathcal{V}_j}
\left(y_i-\hat{f}^{(-j)}(x_i)\right)^2\right).
$$

When $k=n$, this becomes LOOCV. Common choices such as $k=5$ or $k=10$
require far fewer fits while still using every observation for validation.
Smaller $k$ can have slightly more bias because each training set is smaller,
but it is usually more computationally efficient.

```{r tenfold-cv}
set.seed(17)
kfold_results <- data.frame(
  Degree = 1:10,
  Tenfold_CV_MSE = NA_real_
)

for (degree in kfold_results$Degree) {
  model_formula <- if (degree == 1) {
    mpg ~ horsepower
  } else {
    mpg ~ poly(horsepower, degree)
  }
  model <- glm(model_formula, data = Auto)
  kfold_results$Tenfold_CV_MSE[degree] <-
    cv.glm(Auto, model, K = 10)$delta[1]
}

kfold_results
```

```{r compare-cv, fig.width=6, fig.height=4}
plot(
  loocv_results$Degree, loocv_results$LOOCV_MSE,
  type = "b", pch = 19, col = "firebrick",
  xlab = "Polynomial degree", ylab = "Estimated test MSE",
  ylim = range(c(loocv_results$LOOCV_MSE, kfold_results$Tenfold_CV_MSE)),
  main = "LOOCV and 10-fold CV"
)
lines(
  kfold_results$Degree, kfold_results$Tenfold_CV_MSE,
  type = "b", pch = 17, col = "steelblue"
)
legend(
  "topright", legend = c("LOOCV", "10-fold CV"),
  col = c("firebrick", "steelblue"), pch = c(19, 17), lty = 1
)
```

The two curves should have the same practical conclusion even though their
individual values differ slightly. Ten-fold CV is often the useful default:
it is much cheaper than LOOCV and generally gives a stable estimate of the
model's test error. Cross-validation estimates prediction error; it does not
prove that one model is the true data-generating model.

# The Bootstrap

## The basic idea

Cross-validation focuses on prediction. The bootstrap focuses on uncertainty
in an estimate. We treat the observed sample as an approximation to the
population and repeatedly draw bootstrap samples of size $n$ **with
replacement**. For a statistic $\hat{\theta}=T(Z_1,\ldots,Z_n)$, the
$b$th bootstrap estimate is

$$
\hat{\theta}^{*b} = T(Z_1^{*b},\ldots,Z_n^{*b}),
\qquad b=1,\ldots,B.
$$

The standard deviation of the bootstrap estimates approximates the standard
error:

$$
\widehat{\operatorname{SE}}_{\mathrm{boot}}(\hat{\theta})
= \sqrt{\frac{1}{B-1}\sum_{b=1}^{B}
\left(\hat{\theta}^{*b}-\overline{\theta^*}\right)^2}.
$$

The phrase *with replacement* is essential. It means some observations may
appear multiple times and some may be absent in a particular bootstrap
sample. That variation imitates the fact that a new random sample would not
contain exactly the same observations as the original one.

## Portfolio allocation

The `Portfolio` data set contains returns for two assets, `X` and `Y`. Suppose
we invest fraction $\alpha$ in asset $X$ and $1-\alpha$ in asset $Y$. The
portfolio return is

$$R_p = \alpha X + (1-\alpha)Y.$$

Minimizing its variance gives

$$
\hat{\alpha} =
\frac{\hat{\sigma}_Y^2-\hat{\sigma}_{XY}}
{\hat{\sigma}_X^2+\hat{\sigma}_Y^2-2\hat{\sigma}_{XY}}.
$$

This formula is easy to calculate on the original data. The uncertainty in
$\hat{\alpha}$ is less obvious, so it is a good bootstrap example.

```{r portfolio-bootstrap}
data(Portfolio)

alpha_function <- function(data, index) {
  X <- data$X[index]
  Y <- data$Y[index]
  (var(Y) - cov(X, Y)) /
    (var(X) + var(Y) - 2 * cov(X, Y))
}

alpha_hat <- alpha_function(Portfolio, seq_len(nrow(Portfolio)))

set.seed(7)
alpha_boot <- boot(Portfolio, alpha_function, R = 1000)
alpha_se <- sd(alpha_boot$t)
alpha_ci <- alpha_hat + c(-1, 1) * 1.96 * alpha_se

data.frame(
  Estimate = alpha_hat,
  Bootstrap_SE = alpha_se,
  CI_lower = alpha_ci[1],
  CI_upper = alpha_ci[2]
)
```

The estimate is the suggested fraction invested in asset $X$, conditional on
the observed sample. The bootstrap standard error says how much this
allocation would typically move across comparable samples. The approximate
normal-theory interval is computed as

$$
\hat{\alpha} \pm 1.96\,\widehat{\operatorname{SE}}_{\mathrm{boot}}.
$$

This interval is an uncertainty statement, not a guarantee that future
returns will lie inside it. It describes uncertainty in the estimated
allocation, while the investment decision itself also requires a risk model
and practical constraints.

```{r portfolio-bootstrap-plot, fig.width=6, fig.height=4}
hist(
  alpha_boot$t, breaks = 30, col = "grey85", border = "white",
  xlab = expression(alpha^"*"),
  main = "Bootstrap distribution of the portfolio weight"
)
abline(v = alpha_hat, col = "firebrick", lwd = 2)
abline(v = alpha_ci, col = "steelblue", lty = 2, lwd = 2)
legend(
  "topright",
  legend = c("Original estimate", "Approx. 95% interval"),
  col = c("firebrick", "steelblue"), lty = c(1, 2), lwd = 2
)
```

## Bootstrap standard errors for regression coefficients

The bootstrap can estimate uncertainty for almost any statistic that can be
computed from a data set. Here the statistic is the vector of coefficients
from a regression of `mpg` on `horsepower`.

```{r coefficient-bootstrap}
coefficient_function <- function(data, index) {
  coef(lm(mpg ~ horsepower, data = data, subset = index))
}

coefficient_hat <- coefficient_function(Auto, seq_len(nrow(Auto)))

set.seed(1)
coefficient_boot <- boot(Auto, coefficient_function, R = 1000)

coefficient_summary <- data.frame(
  Term = names(coefficient_hat),
  Estimate = unname(coefficient_hat),
  Bootstrap_SE = apply(coefficient_boot$t, 2, sd),
  Classical_SE = coef(summary(lm(mpg ~ horsepower, data = Auto)))[, "Std. Error"]
)
coefficient_summary
```

The slope is negative: cars with greater horsepower tend to have lower fuel
efficiency. The bootstrap standard error measures the variability of that
slope under repeated samples. Comparing it with the usual regression
standard error is a useful diagnostic. Similar values suggest that the
classical approximation is reasonable here; noticeable differences remind us
that standard errors depend on assumptions about the model and its errors.

We can also bootstrap a nonlinear specification. The quadratic coefficient
measures curvature, so its uncertainty helps us decide whether the observed
curvature is stable enough to support the added model complexity.

```{r quadratic-coefficient-bootstrap}
quadratic_function <- function(data, index) {
  coef(lm(mpg ~ horsepower + I(horsepower^2),
          data = data, subset = index))
}

set.seed(1)
quadratic_boot <- boot(Auto, quadratic_function, R = 1000)

data.frame(
  Term = names(coef(lm(mpg ~ horsepower + I(horsepower^2), data = Auto))),
  Estimate = coef(lm(mpg ~ horsepower + I(horsepower^2), data = Auto)),
  Bootstrap_SE = apply(quadratic_boot$t, 2, sd)
)
```

# Putting the Methods Together

The methods answer different questions, so their outputs should not be
confused:

| Method | Main question | Mechanism |
|---|---|---|
| Validation set | How well does this model predict on one held-out split? | One train/validation split |
| LOOCV | What is the average held-out error when each observation is left out? | $n$ fits |
| $k$-fold CV | What is held-out error using a practical number of folds? | $k$ fits |
| Bootstrap | How variable is an estimate across plausible samples? | Resample with replacement |

For the `Auto` analysis, the resampling results generally support a
quadratic relationship between `mpg` and `horsepower`: degree two improves on
degree one, while higher degrees provide little dependable improvement. This
is a model-selection conclusion about predictive performance, not a claim
that a quadratic law is physically exact.

The broader lesson is that a fitted model is only one estimate obtained from
one sample. Validation and cross-validation expose how prediction changes
when observations are held out. The bootstrap exposes how statistical
summaries change when the observed sample is perturbed. Together they provide
an empirical way to reason about generalization and uncertainty rather than
reporting training fit alone.

# Conclusion

Resampling methods turn a single data set into a collection of informative
experiments. The validation set is simple but unstable. LOOCV uses data
efficiently but can be expensive. Ten-fold cross-validation is often a good
practical compromise for comparing model complexity. The bootstrap is more
general: it estimates standard errors for statistics whose sampling
distributions are difficult to derive analytically.

The most important habit is to match the resampling method to the question.
Use held-out predictions when the concern is generalization, and use
bootstrap replicates when the concern is the uncertainty of an estimate.
Neither method removes irreducible noise, but both make the limits of what we
learn from a finite sample much more visible.