---
title: "Lab - Linear Regression"
author: "Wayne Rubangisa"
date: "2026-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(MASS)
# library(ISLR2) We need to install this first in Packages.

# Literally by running the command: 
# install.packages("ISLR2")
```

```{r}
library(ISLR2)
```

To access the Boston housing data-set from the MASS package, use the attach() function


```{r}
attach(Boston)
# Please remember to detach it. 
head(Boston)
```
```{r}
names(Boston)
```


Now, lets start with a simple linear model with `lm()`


```{r}
lm.fit <- lm(medv ~ lstat)
```

Instead of attaching the Boston data-set first, we can also specify it directly in `lm()`.


```{r}
lm.fit <- lm(medv ~ lstat, data = Boston)
```

We can examine the fitted model, by simply typing its name.


```{r}
lm.fit
```
The names() function lists the names of all variables store in the lm.fit model.

```{r}
names(lm.fit)
```


```{r}
coef(lm.fit)
```


```{r}
confint(lm.fit)
```

We use the predict() function to obtain prediction interval or confidence intervals for a given value of `lstat`


```{r}
predict(lm.fit, data.frame(lstat = (c(5, 10, 15))), interval = "confidence")
```

```{r}
predict(lm.fit, data.frame(lstat = (c(5, 10, 15))), interval = "prediction")
```


We can plot the variables `lstat` and `medv` using the plot() function, and overlay a regression line from our linear model using the abline() function.

```{r}
plot(lstat, medv)
abline(lm.fit)
```


Experiment with different options for abline() by changing the line width and color.

```{r}
plot(lstat, medv)
abline(lm.fit, lwd = 3)
```




```{r}
plot(lstat, medv)
abline(lm.fit, lwd = 3, col = "red")
```


```{r}
plot(lstat, medv, col = "red")
```


```{r}
plot(lstat, medv, pch = 20)
```

```{r}
plot(lstat, medv, pch = "+")
```


```{r}
plot(1:20, 1:20, pch = 1:20)
```

To create a grid of multiple subplots, we'll use the par() function.

```{r}
par(mfrow = c(2, 2))
plot(lm.fit)
```


We can extract the residuals and studentized residuals from our linear model using `residuals()` and `rstudent()`, and plot them along with the predicted values.

```{r}
plot(predict(lm.fit), residuals(lm.fit))
```

```{r}
plot(predict(lm.fit), rstudent(lm.fit))
```


```{r}
plot(hatvalues(lm.fit))
```

```{r}
which.max(hatvalues(lm.fit))
```


## 3.6.3 Multiple Linear Regression

The `lm()` can be used for fitting multiple regression models as well. In the following example, we include age in addition to `lstata` as predictors of of our response variable `medv`.

```{r}
lm.fit <- lm(medv ~ lstat + age, data = Boston)
summary(lm.fit)
```

In a formula, a dot `.` can be used as a shorthand to include all variables from the Boston data-set as predictors.

```{r}
lm.fit <- lm(medv ~ ., data = Boston)
summary(lm.fit)
```

The Variance Inflation Factors (VIF) can be calculated using the `vif()` function from the `Companion to Applied Regression` or `car` package. Let's install it using `install.packages()` 


```{r}
library(car)
vif(lm.fit)
```


To exclude specific variables from the list of predictors, we can use the - notation. In the following example, all variables except `age` are included in the linear fit model.

```{r}
lm.fit1 <- lm(medv ~ . - age, data = Boston)
summary(lm.fit1)
```
We can also update an existing fit with the update() function and specifying a new formula.

```{r}
lm.fit1 <- update(lm.fit, ~. - age)
```

## 3.6.4 Interaction Terms


```{r}
summary(lm(medv ~ lstat * age, data = Boston))
```

## 3.6.5 Non-linear Transformations of the Predictors

The `lm()` function can also be non-linear transformations such as quadratic or cubic terms. Terms in the formula raised to a power with `^` must be wrapped inside a call to `I()` to prevent interpretation and treated `as is`.

```{r}
lm.fit2 <- lm(medv ~ lstat + I(lstat^2))
summary(lm.fit2)
```

To examine the analysis of variance (ANOVA) for one or more models, we use the anova().

```{r}
lm.fit <- lm(medv ~ lstat)
anova(lm.fit, lm.fit2)
```


```{r}
par(mfrow = c(2, 2))
plot(lm.fit2)
```


```{r}
lm.fit5 <- lm(medv ~ poly(lstat, 5))
summary(lm.fit5)
```

## 3.6.6 Qualitative Predictors

```{r}
head(Carseats)
```


```{r}
names(Carseats)
```

```{r}
lm.fit <- lm(Sales ~ . + Income:Advertising + Price:Age, data = Carseats)
summary(lm.fit)
```

To examine the coding for the qualitative variables, we can use the contrasts() function.

```{r}
contrasts(Carseats$ShelveLoc)
```


## 3.6.7 Writing Functions

We can define our own functions to wrap a set of R commands in a single call.

```{r}
LoadLibraries <- function() {
    library(ISLR)
    library(MASS)
}
```















































