---
output:
  pdf_document:
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{amsmath}
  - \input{../../_common/preamble.tex}
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
```

```{=latex}
\makecoverpage{Chapter 4 --- Classification \\ Exercises 5, 10, 14, and 16}
```

```{r libraries}
library(ISLR2)
library(MASS)
library(class)
library(e1071)
```

Note: this document loads both `ISLR2` and `MASS`, and both packages
export a data set called `Boston` (`MASS::Boston` has 14 columns,
including a variable dropped from `ISLR2::Boston`'s 13). Because `MASS`
is attached after `ISLR2`, it sits earlier on R's search path — so an
unqualified `Boston` would silently resolve to `MASS::Boston`, not the
one the book actually uses. Exercise 16 below refers to it explicitly as
`ISLR2::Boston` throughout to avoid that trap.

# Exercise 5 — LDA vs. QDA

## (a) Bayes boundary is linear

**Training set:** QDA will (weakly) achieve better training performance —
it has strictly more flexibility, so it can always fit the training data at
least as well as LDA, and typically better, by using its extra covariance
parameters to bend the boundary even when it doesn't need to.

**Test set:** LDA should perform better. If the true Bayes boundary really
is linear, LDA's assumption matches reality exactly, while QDA is
estimating a more flexible boundary than the data actually calls for.
That extra flexibility doesn't reduce bias here (there's no bias to
reduce — LDA is already unbiased for a linear boundary), so it's pure
added variance from estimating a much larger number of covariance
parameters, which typically shows up as worse test performance.

## (b) Bayes boundary is non-linear

**Training set:** QDA again fits at least as well, for the same reason as
(a).

**Test set:** QDA should perform better. Now LDA's linear assumption is
genuinely wrong, so it carries irreducible bias no amount of data can fix.
QDA can actually approximate the curvature, trading some added variance
for a real reduction in bias — and as long as there's enough data to
estimate the extra parameters reasonably well, that trade tends to pay off.

## (c) As $n$ increases

We'd expect QDA's relative test accuracy to **improve**. QDA has more
parameters to estimate (a full covariance matrix per class, rather than
one shared matrix), so its variance is more sensitive to sample size.
With more data, QDA's parameter estimates get more stable — closing the
variance gap with LDA — while gaining nothing changes for LDA's
bias if the truth isn't linear. So more data specifically helps the
model that needed it most.

## (d) True or False: QDA is superior even when the Bayes boundary is linear, because it's flexible enough to approximate a linear boundary

**False.** This is exactly the setup from part (a). Yes, QDA is flexible
enough to approximate a linear boundary reasonably closely — but
"flexible enough to approximate it" is not the same as "better at
estimating it." QDA has to *estimate* that near-linear shape from finite,
noisy data using far more parameters than LDA needs, and that estimation
variance is real cost with no offsetting benefit, since LDA already
targets the right boundary shape exactly. The claim mistakes flexibility
for accuracy — more flexibility only helps when the extra bias reduction
it buys outweighs the extra variance it costs, and here there's no bias to
reduce in the first place.

\newpage

# Exercise 10 — The LDA Log-Odds Formula for $p = 1$

The general ($p > 1$) result gives the log-odds between class $k$ and a
baseline class $K$ as a linear function of $x$:

$$
\log\left(\frac{\Pr(Y=k \mid X=x)}{\Pr(Y=K \mid X=x)}\right) = a_k + \sum_{j=1}^p b_{kj} x_j
$$

with $a_k$ and $b_{kj}$ expressed in terms of the class priors $\pi_k$,
class mean vectors $\mu_k$, and the shared covariance matrix $\Sigma$. For
$p=1$, every one of those quantities collapses to a scalar: $\mu_k$
becomes a single number, and $\Sigma$ becomes the shared variance
$\sigma^2$.

Under LDA's assumptions, $X \mid Y=k \sim N(\mu_k, \sigma^2)$, so the
one-dimensional normal density is

$$
f_k(x) = \frac{1}{\sqrt{2\pi}\,\sigma} \exp\left(-\frac{(x-\mu_k)^2}{2\sigma^2}\right)
$$

By Bayes' theorem, $\Pr(Y=k\mid X=x) \propto \pi_k f_k(x)$, so the log-odds
between class $k$ and class $K$ is

$$
\log\left(\frac{\Pr(Y=k\mid X=x)}{\Pr(Y=K\mid X=x)}\right)
= \log\left(\frac{\pi_k f_k(x)}{\pi_K f_K(x)}\right)
= \log\left(\frac{\pi_k}{\pi_K}\right) + \log\left(\frac{f_k(x)}{f_K(x)}\right)
$$

The $\frac{1}{\sqrt{2\pi}\sigma}$ normalizing constants are identical for
every class (shared $\sigma^2$), so they cancel in the ratio $f_k(x)/f_K(x)$,
leaving

$$
\log\left(\frac{f_k(x)}{f_K(x)}\right)
= -\frac{(x-\mu_k)^2}{2\sigma^2} + \frac{(x-\mu_K)^2}{2\sigma^2}
= \frac{1}{2\sigma^2}\Big[(x-\mu_K)^2 - (x-\mu_k)^2\Big]
$$

Expanding both squares — $(x-\mu_K)^2 = x^2 - 2x\mu_K + \mu_K^2$ and
similarly for $\mu_k$ — the $x^2$ terms cancel between them, leaving only
terms linear in $x$ plus constants:

$$
(x-\mu_K)^2 - (x-\mu_k)^2 = 2x(\mu_k - \mu_K) - (\mu_k^2 - \mu_K^2)
$$

Putting it all together:

$$
\log\left(\frac{\Pr(Y=k\mid X=x)}{\Pr(Y=K\mid X=x)}\right)
= \underbrace{\log\left(\frac{\pi_k}{\pi_K}\right) - \frac{\mu_k^2 - \mu_K^2}{2\sigma^2}}_{a_k}
\; + \; \underbrace{\frac{\mu_k - \mu_K}{\sigma^2}}_{b_{k1}} \cdot x
$$

so, matching the target form (with the single predictor playing the role
of $x_1$, i.e. $j=1$):

$$
a_k = \log\left(\frac{\pi_k}{\pi_K}\right) - \frac{\mu_k^2 - \mu_K^2}{2\sigma^2},
\qquad
b_{k1} = \frac{\mu_k - \mu_K}{\sigma^2}
$$

This is exactly the general $p$-dimensional result specialized to a single
feature: $a_k$ still combines the log prior-odds with a term penalizing
classes whose mean is far from the origin (relative to their shared
variance), and $b_{k1}$ is still $\Sigma^{-1}(\mu_k - \mu_K)$ — just with
matrix inversion trivialized to ordinary division by the scalar $\sigma^2$.

\newpage

# Exercise 14 — Predicting High/Low `mpg` for the `Auto` Data

## (a) Creating `mpg01`

```{r mpg01}
Auto2 <- Auto
Auto2$mpg01 <- as.integer(Auto2$mpg > median(Auto2$mpg))
table(Auto2$mpg01)
```

Splitting exactly at the median naturally gives a near-perfectly balanced
binary outcome (196/196) — useful, since classification methods generally
work best when neither class dominates.

## (b) Exploring associations with `mpg01`

```{r mpg01-explore, fig.width=7, fig.height=5}
par(mfrow = c(2, 2))
boxplot(weight ~ mpg01, data = Auto2, main = "Weight vs mpg01",
        xlab = "mpg01", ylab = "weight")
boxplot(displacement ~ mpg01, data = Auto2, main = "Displacement vs mpg01",
        xlab = "mpg01", ylab = "displacement")
boxplot(horsepower ~ mpg01, data = Auto2, main = "Horsepower vs mpg01",
        xlab = "mpg01", ylab = "horsepower")
boxplot(cylinders ~ mpg01, data = Auto2, main = "Cylinders vs mpg01",
        xlab = "mpg01", ylab = "cylinders")
```

```{r mpg01-corr}
sort(abs(sapply(Auto2[, c("cylinders","displacement","horsepower","weight",
                            "acceleration","year","origin")],
                 function(x) cor(x, Auto2$mpg01))), decreasing = TRUE)
```

`weight`, `displacement`, `horsepower`, and `cylinders` all separate the
two `mpg01` groups cleanly in the boxplots and are the four most
correlated with `mpg01` (all above 0.75 in magnitude) — heavier, larger-
engined, more-cylindered cars cluster in the low-`mpg` group, which
matches basic physical intuition (more mass and engine displacement to
move costs more fuel per mile). These four are used as predictors below.

## (c) Train/test split

```{r mpg01-split}
set.seed(1)
n <- nrow(Auto2)
train <- sample(seq_len(n), size = floor(0.7 * n))
train.set <- Auto2[train, ]
test.set <- Auto2[-train, ]

vars <- c("cylinders", "displacement", "horsepower", "weight")
form <- as.formula(paste("mpg01 ~", paste(vars, collapse = " + ")))
c(train = nrow(train.set), test = nrow(test.set))
```

## (d) LDA

```{r mpg01-lda}
lda.fit <- lda(form, data = train.set)
lda.pred <- predict(lda.fit, test.set)$class
lda.err <- mean(lda.pred != test.set$mpg01)
lda.err
```

## (e) QDA

```{r mpg01-qda}
qda.fit <- qda(form, data = train.set)
qda.pred <- predict(qda.fit, test.set)$class
qda.err <- mean(qda.pred != test.set$mpg01)
qda.err
```

## (f) Logistic regression

```{r mpg01-logistic}
glm.fit <- glm(form, data = train.set, family = binomial)
glm.probs <- predict(glm.fit, test.set, type = "response")
glm.pred <- ifelse(glm.probs > 0.5, 1, 0)
glm.err <- mean(glm.pred != test.set$mpg01)
glm.err
```

## (g) Naive Bayes

```{r mpg01-nb}
nb.fit <- naiveBayes(form, data = train.set)
nb.pred <- predict(nb.fit, test.set)
nb.err <- mean(nb.pred != test.set$mpg01)
nb.err
```

## (h) KNN across several values of $K$

```{r mpg01-knn}
train.X <- scale(train.set[, vars])
test.X <- scale(test.set[, vars],
                 center = attr(train.X, "scaled:center"),
                 scale = attr(train.X, "scaled:scale"))
train.Y <- train.set$mpg01

k.values <- c(1, 3, 5, 7, 9, 11, 15, 21)
knn.errs <- sapply(k.values, function(k) {
  set.seed(1)
  pred <- knn(train.X, test.X, train.Y, k = k)
  mean(pred != test.set$mpg01)
})
data.frame(k = k.values, test_error = knn.errs)
```

```{r mpg01-summary}
results <- data.frame(
  method = c("LDA", "QDA", "Logistic", "Naive Bayes",
             paste0("KNN (k=", k.values, ")")),
  test_error = c(lda.err, qda.err, glm.err, nb.err, knn.errs)
)
results[order(results$test_error), ]
```

**Logistic regression gives the lowest test error** among all methods
tried, with naive Bayes and the better-performing KNN settings
(k = 11, 15, 21) close behind, and LDA/QDA tied and slightly worse. KNN
with very small $k$ (1, 5, 9) is the *worst* performer here — consistent
with what we saw in the lab: an overly flexible, low-$k$ neighbor vote is
prone to fitting noise in only 274 training points across four
predictors, and needs a larger neighborhood to stabilize. With
well-separated, roughly-Gaussian-looking predictors like these
(supported by the boxplots in (b)), it's unsurprising that a
straightforward, low-variance method like logistic regression comes out
ahead of the more flexible alternatives.

\newpage

# Exercise 16 — Predicting High/Low Crime Rate for the `Boston` Data

```{r crim01-setup}
Boston2 <- ISLR2::Boston
Boston2$crim01 <- as.integer(Boston2$crim > median(Boston2$crim))
```

We create `crim01` ourselves, exactly as `mpg01` was created in Exercise
14, and use the same logic to choose predictors: check which variables
correlate most strongly with the new binary outcome.

```{r crim01-corr}
preds <- setdiff(names(Boston2), c("crim", "crim01"))
corrs <- sort(abs(sapply(Boston2[, preds], function(x) cor(x, Boston2$crim01))),
              decreasing = TRUE)
round(corrs, 3)
```

`nox`, `rad`, `dis`, `age`, `tax`, and `indus` all correlate with
`crim01` above 0.6 in magnitude — tracts with more industrial land use,
worse air quality, easier highway access, older housing, and higher
property tax all tend to have above-median crime, echoing the same
pattern found back in Chapter 2's exploration of this data set. These six
are used as predictors below.

```{r crim01-split}
set.seed(1)
n <- nrow(Boston2)
train <- sample(seq_len(n), size = floor(0.7 * n))
train.set <- Boston2[train, ]
test.set <- Boston2[-train, ]

vars <- c("nox", "rad", "dis", "age", "tax", "indus")
form <- as.formula(paste("crim01 ~", paste(vars, collapse = " + ")))
```

## Logistic regression

```{r crim01-logistic}
glm.fit <- glm(form, data = train.set, family = binomial)
glm.probs <- predict(glm.fit, test.set, type = "response")
glm.pred <- ifelse(glm.probs > 0.5, 1, 0)
glm.err <- mean(glm.pred != test.set$crim01)
glm.err
```

## LDA

```{r crim01-lda}
lda.fit <- lda(form, data = train.set)
lda.pred <- predict(lda.fit, test.set)$class
lda.err <- mean(lda.pred != test.set$crim01)
lda.err
```

## Naive Bayes

```{r crim01-nb}
nb.fit <- naiveBayes(form, data = train.set)
nb.pred <- predict(nb.fit, test.set)
nb.err <- mean(nb.pred != test.set$crim01)
nb.err
```

## KNN across several values of $K$

```{r crim01-knn}
train.X <- scale(train.set[, vars])
test.X <- scale(test.set[, vars],
                 center = attr(train.X, "scaled:center"),
                 scale = attr(train.X, "scaled:scale"))
train.Y <- train.set$crim01

k.values <- c(1, 3, 5, 7, 9, 11)
knn.errs <- sapply(k.values, function(k) {
  set.seed(1)
  pred <- knn(train.X, test.X, train.Y, k = k)
  mean(pred != test.set$crim01)
})
data.frame(k = k.values, test_error = knn.errs)
```

## Findings

```{r crim01-summary}
results <- data.frame(
  method = c("Logistic", "LDA", "Naive Bayes", paste0("KNN (k=", k.values, ")")),
  test_error = c(glm.err, lda.err, nb.err, knn.errs)
)
results[order(results$test_error), ]
```

**KNN clearly wins here** (best around k = 3–7, roughly a third the error
rate of the parametric methods), with logistic regression next, LDA
close behind, and naive Bayes distinctly worst. This is a genuinely
different ranking from Exercise 14 — worth pausing on why. `nox`, `rad`,
`dis`, `age`, `tax`, and `indus` are far from independent of each other
(recall the `tax`/`rad` clustering pattern from Chapter 2's pairwise
scatterplots), which directly undercuts naive Bayes's core assumption and
explains why it performs worst. And crucially, the true relationship
between these predictors and high/low crime plausibly isn't well
approximated by any smooth linear or Gaussian-boundary form at all — the
Chapter 2 exploration found crime concentrated in a distinct cluster of
tracts with extreme `tax`/`rad` values rather than varying smoothly with
these predictors, which is exactly the kind of localized, irregular
structure a flexible, non-parametric method like KNN can capture and a
linear-boundary method (logistic regression, LDA) or an
independence-assuming method (naive Bayes) cannot.
