---
output:
  pdf_document:
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{amsmath}
  - \input{../../_common/preamble.tex}
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

```{=latex}
\makecoverpage{Chapter 8 --- Tree-Based Methods \\ Lab}
```

# Introduction

A decision tree predicts an outcome by repeatedly splitting the predictor
space into simpler regions. The result is easy to explain: observations that
follow the same sequence of splits receive the same prediction. Trees are
therefore attractive when interpretability, nonlinear relationships, and
interactions matter.

Their main weakness is instability. A small change in the training data can
change an early split, which can change the entire tree below it. This lab
starts with individual classification and regression trees, then introduces
three ensemble approaches that address this instability in different ways:

1. pruning controls the complexity of one tree;
2. Bagging averages many trees trained on bootstrap samples;
3. Random Forests add random feature selection to reduce correlation between
   the trees; and
4. Boosting builds trees sequentially to correct the current model's errors.

The lab follows the Chapter 8.3 workflow from *An Introduction to Statistical
Learning*. The explanations state the practical idea first and then connect it
to the relevant statistical objective.

```{r libraries}
# Install these once if necessary:
# install.packages(c("ISLR2", "tree", "randomForest", "gbm", "BART"))

library(ISLR2)
library(tree)
library(randomForest)
library(gbm)
library(BART)
```

# Classification Trees

## The `Carseats` data

The `Carseats` data describe the sales of child car seats at 400 stores. Our
first task is classification: we create a binary response indicating whether
sales are high or low.

A store is labelled `Yes` when `Sales` exceeds 8 thousand units and `No`
otherwise. This threshold is a modelling choice: it changes a continuous
business outcome into a decision problem that could support an action such as
identifying stores for extra promotion.

```{r carseats-prepare}
data(Carseats)
High <- factor(ifelse(Carseats$Sales <= 8, "No", "Yes"))
Carseats <- data.frame(Carseats, High)

str(Carseats)
table(Carseats$High)
```

## Growing a classification tree

A classification tree partitions the predictor space into regions and assigns
one class to each terminal region. At a candidate split, the algorithm looks
for a partition that makes the resulting child nodes purer. One common purity
measure is the Gini index:

$$
G = \sum_{k=1}^{K} \hat p_k(1-\hat p_k),
$$

where $\hat p_k$ is the proportion of observations in the node belonging to
class $k$. A node containing only one class has Gini index zero; a mixed node
has a larger value.

The `tree()` function grows a large tree first. This is useful because the
subsequent pruning step can compare smaller subtrees obtained from the large
candidate tree.

```{r carseats-tree}
tree.carseats <- tree(High ~ . - Sales, data = Carseats)
tree.carseats

cat("Terminal nodes:", sum(tree.carseats$frame$var == "<leaf>"), "\n")
cat("Training misclassification rate:",
  mean(predict(tree.carseats, type = "class") != Carseats$High), "\n")
```

The summary reports the number of terminal nodes and the training error. A
very large tree can have a low training error because it has enough regions to
memorise the training sample. That is not evidence that it will predict new
stores well.

```{r carseats-tree-plot, fig.width=8, fig.height=6}
plot(tree.carseats)
text(tree.carseats, pretty = 0)
```

The plot should be read from the root downward. At each internal node, the
condition sends an observation left or right. A terminal node gives the final
class prediction and its estimated class probabilities.

## Cross-validation and pruning

A tree with too many terminal nodes has high variance, while a tree that is
too small may have high bias. Cost-complexity pruning makes this tradeoff
explicit. For a subtree $T$, a common objective is

$$
R_\alpha(T) = R(T) + \alpha |T|,
$$

where $R(T)$ measures the tree's training misclassification error, $|T|$ is
the number of terminal nodes, and $\alpha \geq 0$ penalizes complexity. A
larger $\alpha$ favours a smaller tree.

Cross-validation estimates which tree size is likely to perform best on new
data. Here we use misclassification error as the cross-validation criterion.

```{r carseats-cv}
set.seed(1)
cv.carseats <- cv.tree(tree.carseats, FUN = prune.misclass)
cv.carseats

plot(cv.carseats$size, cv.carseats$dev, type = "b",
     xlab = "Number of terminal nodes",
     ylab = "Cross-validated misclassification error",
     main = "Choosing the classification-tree size")
```

The lowest point is the cross-validation choice. Because cross-validation is
random, a different seed can move the minimum slightly; the important idea is
that tree size is selected using held-out predictions rather than training
error alone.

```{r carseats-prune}
prune.carseats <- prune.misclass(tree.carseats,
                                 best = cv.carseats$size[which.min(cv.carseats$dev)])
summary(prune.carseats)

plot(prune.carseats)
text(prune.carseats, pretty = 0)
```

Pruning usually produces a simpler tree. Simplicity is useful for
interpretation, but it is not automatically optimal; the cross-validation
estimate is the evidence used to choose the amount of pruning.

# Regression Trees

## The `Hitters` data

We now predict a quantitative response: baseball players' salaries. The
regression-tree analogue of a classification tree assigns a constant
prediction to every terminal node, usually the mean response of the training
observations in that node.

A regression tree chooses splits to reduce residual sum of squares (RSS):

$$
\operatorname{RSS}(T) = \sum_{m=1}^{|T|}
  \sum_{i:x_i \in R_m}(y_i-\hat y_{R_m})^2,
$$

where $R_m$ is a terminal region and $\hat y_{R_m}$ is the mean response in
that region. Again, an extremely large tree can reduce training RSS while
performing poorly on new observations.

```{r hitters-prepare}
data(Hitters)
Hitters <- na.omit(Hitters)

set.seed(1)
train <- sample(1:nrow(Hitters), nrow(Hitters) / 2)
Hitters.train <- Hitters[train, ]
Hitters.test <- Hitters[-train, ]
```

```{r hitters-tree}
tree.hitters <- tree(Salary ~ ., data = Hitters.train)
summary(tree.hitters)

plot(tree.hitters)
text(tree.hitters, pretty = 0)
```

The first split is the predictor that gives the largest immediate reduction in
RSS under the algorithm's greedy search. Later splits represent interactions:
their meaning is conditional on the decisions made higher in the tree.

## Cross-validation and a test-set check

```{r hitters-cv}
set.seed(1)
cv.hitters <- cv.tree(tree.hitters)
plot(cv.hitters$size, cv.hitters$dev, type = "b",
     xlab = "Number of terminal nodes",
     ylab = "Cross-validated RSS",
     main = "Choosing the regression-tree size")

prune.hitters <- prune.tree(tree.hitters,
                            best = cv.hitters$size[which.min(cv.hitters$dev)])
summary(prune.hitters)
```

We evaluate the unpruned and pruned models on the same test set. Test error is
an estimate of future prediction error because these observations were not
used to fit the tree.

```{r hitters-test-error}
pred.tree <- predict(tree.hitters, Hitters.test)
pred.pruned <- predict(prune.hitters, Hitters.test)

mse.tree <- mean((pred.tree - Hitters.test$Salary)^2)
mse.pruned <- mean((pred.pruned - Hitters.test$Salary)^2)

c(unpruned_MSE = mse.tree, pruned_MSE = mse.pruned)
```

The better model is the one with the smaller test MSE in this split. The
comparison also illustrates an important qualification: pruning can improve
generalisation, but the result depends on the sample split and on the signal
in the data. The purpose of the test set is to assess the final choice, not to
choose repeatedly until a favourable result appears.

# Bagging and Random Forests

## Why averaging trees helps

Bagging means bootstrap aggregating. We repeatedly sample $n$ observations
with replacement, grow a tree on each bootstrap sample, and average the
predictions. For regression,

$$
\hat f_{\mathrm{bag}}(x) = \frac{1}{B}\sum_{b=1}^{B}T_b(x).
$$

If the trees have variance $\sigma^2$ and pairwise correlation $\rho$, then
under the usual simplifying assumptions,

$$
\operatorname{Var}\left(\frac{1}{B}\sum_{b=1}^{B}T_b(x)\right)
= \sigma^2\left(\rho + \frac{1-\rho}{B}\right).
$$

Increasing $B$ reduces the unshared part of the variance. It cannot remove
error shared by all trees, which is why reducing tree correlation is valuable.

## Bagging on `Boston`

The textbook uses the `Boston` data to compare tree ensembles for predicting
`medv`, the median value of owner-occupied homes in a Boston suburb. We use a
fixed train/test split so the models are compared on exactly the same cases.

```{r boston-prepare}
Boston <- ISLR2::Boston
set.seed(1)
train <- sample(1:nrow(Boston), nrow(Boston) / 2)
Boston.train <- Boston[train, ]
Boston.test <- Boston[-train, ]
```

For bagging, `mtry = 13` allows each split to consider all predictors. The
bootstrap samples create the variation between trees.

```{r boston-bagging}
set.seed(1)
bag.boston <- randomForest(
  medv ~ ., data = Boston.train,
  mtry = ncol(Boston.train) - 1,
  importance = TRUE
)

bag.pred <- predict(bag.boston, Boston.test)
bag.mse <- mean((bag.pred - Boston.test$medv)^2)
bag.mse
```

The out-of-bag estimate is produced by predicting each training observation
using only trees whose bootstrap sample did not contain that observation. It
provides an internal performance estimate without a separate validation set.

```{r boston-bagging-oob}
bag.boston$mse[length(bag.boston$mse)]
```

## Random Forests

Random Forests add a second source of randomness. At each split, only a
random subset of the predictors is considered. If the strongest predictor is
not always available, different trees are forced to explore different
structures, reducing their correlation $\rho$.

```{r boston-random-forest}
set.seed(1)
rf.boston <- randomForest(
  medv ~ ., data = Boston.train,
  mtry = 6,
  importance = TRUE
)

rf.pred <- predict(rf.boston, Boston.test)
rf.mse <- mean((rf.pred - Boston.test$medv)^2)
rf.mse

importance(rf.boston)
varImpPlot(rf.boston)
```

The test MSEs provide an empirical comparison of bagging and the Random
Forest. The Random Forest is not guaranteed to win on every split: its design
trades a little individual-tree strength for lower correlation, and the
tradeoff is data-dependent.

```{r boston-ensemble-summary}
data.frame(
  Method = c("Bagging", "Random Forest"),
  Test_MSE = c(bag.mse, rf.mse)
)
```

# Boosting

## Sequential error correction

Boosting does not grow independent trees and average them. Instead, it builds
an additive model one tree at a time:

$$
F_m(x) = F_{m-1}(x) + \lambda h_m(x),
$$

where $h_m$ is a new, usually shallow, tree and $\lambda$ is the learning
rate. For squared-error regression, the next tree is fitted to the current
residuals $y_i-F_{m-1}(x_i)$. The new tree is therefore a correction, not a
complete model trained from scratch.

## Gradient boosting on `Boston`

The `gbm()` implementation uses stochastic gradient boosting. Its important
controls are the number of trees, interaction depth, and shrinkage (learning
rate). A smaller learning rate usually requires more trees but makes the
updates more gradual.

```{r boston-boosting}
set.seed(1)
boost.boston <- gbm(
  medv ~ ., data = Boston.train,
  distribution = "gaussian",
  n.trees = 5000,
  interaction.depth = 4,
  shrinkage = 0.01,
  verbose = FALSE
)

summary(boost.boston)
```

The variable-importance plot summarizes how much each variable contributes to
reducing the loss across the boosted trees. It is a useful descriptive tool,
not proof that a variable causes the response.

```{r boston-boosting-importance, fig.width=7, fig.height=5}
par(mar = c(5, 10, 4, 2))
summary(boost.boston, las = 1)
```

```{r boston-boosting-mse}
boost.pred <- predict(boost.boston, Boston.test, n.trees = 5000)
boost.mse <- mean((boost.pred - Boston.test$medv)^2)
boost.mse

ensemble.results <- data.frame(
  Method = c("Bagging", "Random Forest", "Boosting"),
  Test_MSE = c(bag.mse, rf.mse, boost.mse)
)
ensemble.results[order(ensemble.results$Test_MSE), ]
```

The smallest test MSE in this table is the best result for this particular
split and set of tuning choices. It should not be interpreted as a universal
ranking. In practice, the number of trees, depth, learning rate, and feature
sampling settings should be selected using cross-validation or a validation
set, and the final test set should be used only once for the final assessment.

# Bayesian Additive Regression Trees

Bayesian Additive Regression Trees (BART) also represent the response as a
sum of many trees:

$$
Y = \sum_{j=1}^{m} g(X; T_j, M_j) + \varepsilon,
\qquad \varepsilon \sim N(0, \sigma^2).
$$

Here $T_j$ is the structure of tree $j$ and $M_j$ contains its terminal-node
values. Unlike ordinary boosting, BART puts prior distributions on tree
structures and leaf parameters, then uses posterior sampling to represent
uncertainty about the ensemble.

```{r boston-bart}
set.seed(1)
x.train <- as.matrix(Boston.train[, setdiff(names(Boston.train), "medv")])
y.train <- Boston.train$medv
x.test <- as.matrix(Boston.test[, setdiff(names(Boston.test), "medv")])

bart.fit <- wbart(
  x.train, y.train,
  x.test = x.test,
  nskip = 1000,
  ndpost = 2000,
  ntree = 200
)

bart.pred <- colMeans(bart.fit$yhat.test)
bart.mse <- mean((bart.pred - Boston.test$medv)^2)
bart.mse
```

The BART prediction is the posterior mean of the test predictions. Its
posterior draws also contain uncertainty information, which is a major
conceptual difference from a single fitted tree or a point-prediction-only
ensemble.

```{r boston-final-summary}
final.results <- rbind(
  ensemble.results,
  data.frame(Method = "BART", Test_MSE = bart.mse)
)
final.results[order(final.results$Test_MSE), ]
```

# Conclusion

A single tree gives an interpretable sequence of decisions, but its structure
can be unstable. Pruning controls that instability by selecting a smaller
subtree. Bagging reduces variance by averaging trees grown on bootstrap
samples. Random Forests improve the averaging strategy by also reducing the
correlation between trees through random feature selection. Boosting follows a
different path: it grows trees sequentially, with each tree correcting the
current ensemble. BART extends the additive-tree idea with Bayesian priors
and posterior uncertainty.

The main lesson is not that one method always wins. The appropriate method
depends on the response type, the signal-to-noise ratio, the amount of tuning
available, and whether interpretability or predictive accuracy is the primary
goal. A fair comparison requires the same data split or resampling scheme,
careful control of tuning, and a final evaluation on observations not used to
fit the chosen model.
