---
output:
  pdf_document:
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{amsmath}
  - \input{../../_common/preamble.tex}
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
```

```{=latex}
\makecoverpage{Chapter 3 --- Linear Regression \\ Lab}
```

# Introduction

This lab walks through fitting and interpreting linear regression models in R,
following the workflow used in Chapter 3 of *An Introduction to Statistical
Learning*. We'll cover:

- Simple linear regression
- Confidence and prediction intervals
- Diagnostic plots for checking model assumptions
- Multiple linear regression
- Interaction terms
- Non-linear transformations of predictors
- Qualitative (categorical) predictors
- Writing a small reusable function

We'll use two classic datasets that ship with the `ISLR2` package: `Boston`
(housing values in the suburbs of Boston) and `Carseats` (simulated child car
seat sales data).

## Required packages

```{r libraries}
# install.packages(c("MASS", "ISLR2", "car"))  # uncomment if needed
library(MASS)     # for some classic datasets and stats utilities
library(ISLR2)    # Boston and Carseats datasets live here (2nd edition)
library(car)      # for vif() -- variance inflation factors
```

> **Note:** In the 2nd edition of ISLR, the `Boston` dataset moved from the
> `MASS` package into `ISLR2`. Loading both packages avoids ambiguity issues;
> if you have both loaded, `ISLR2::Boston` is the one used below.

# Simple Linear Regression

We start by exploring the relationship between `medv` (median home value, in
$1000s) and `lstat` (percentage of the population with "lower status", e.g.
lower income/education) in the `Boston` data set.

```{r explore-data}
Boston <- ISLR2::Boston
str(Boston)
names(Boston)
```

## Fitting the model

Before fitting anything, it's worth being explicit about what `lm()` is
actually estimating. A simple linear regression assumes the true
relationship has the form

$$
\text{medv} = \beta_0 + \beta_1 \cdot \text{lstat} + \epsilon
$$

where $\beta_0$ and $\beta_1$ are unknown population parameters and
$\epsilon$ is irreducible error. `lm()` doesn't know $\beta_0$ and
$\beta_1$ — it estimates them, $\hat\beta_0$ and $\hat\beta_1$, by finding
the line that minimizes the residual sum of squares over the $n$ observed
pairs $(x_i, y_i)$:

$$
\text{RSS} = \sum_{i=1}^{n} \left(y_i - \hat\beta_0 - \hat\beta_1 x_i\right)^2
$$

which has the closed-form (least-squares) solution

$$
\hat\beta_1 = \frac{\sum_{i=1}^n (x_i - \bar{x})(y_i - \bar{y})}{\sum_{i=1}^n (x_i - \bar{x})^2}, \qquad
\hat\beta_0 = \bar{y} - \hat\beta_1 \bar{x}
$$

```{r simple-lm}
lm.fit <- lm(medv ~ lstat, data = Boston)
lm.fit
summary(lm.fit)
```

The output above gives us $\hat\beta_0$ and $\hat\beta_1$ (under
`Estimate`), their standard errors, t-statistics, and p-values, along with
two summary measures of overall fit:

$$
R^2 = 1 - \frac{\text{RSS}}{\text{TSS}}, \qquad
\text{RSE} = \sqrt{\frac{\text{RSS}}{n-2}}
$$

$R^2$ is the proportion of variance in `medv` explained by `lstat` (unitless,
between 0 and 1); the residual standard error (RSE) is roughly the average
distance, in `medv`'s own units (\$1000s), that an observation falls from the
fitted line.

We can pull out specific pieces of the fitted model:

```{r extract-info}
names(lm.fit)
coef(lm.fit)
confint(lm.fit)
```

## Confidence and prediction intervals

Given a value of `lstat`, we can produce a confidence interval for the mean
response and a prediction interval for an individual observation. Both are
centered on the same point prediction $\hat{y}_0 = \hat\beta_0 + \hat\beta_1 x_0$,
but differ in what uncertainty they account for:

$$
\text{CI: } \hat{y}_0 \pm t_{\alpha/2, \, n-2} \cdot \text{SE}(\hat{y}_0), \qquad
\text{PI: } \hat{y}_0 \pm t_{\alpha/2, \, n-2} \cdot \sqrt{\text{SE}(\hat{y}_0)^2 + \hat\sigma^2}
$$

The confidence interval only accounts for uncertainty in *estimating* the
regression line itself (i.e. in $\hat\beta_0, \hat\beta_1$). The prediction
interval adds the irreducible error variance $\hat\sigma^2$ on top of that,
since an individual observation also deviates randomly from the true line —
which is exactly why, as the formulas above make clear, the PI is always
wider than the CI at the same $x_0$.

```{r ci-pi}
new_lstat <- data.frame(lstat = c(5, 10, 15))

predict(lm.fit, new_lstat, interval = "confidence")
predict(lm.fit, new_lstat, interval = "prediction")
```

Note that the prediction interval is wider than the confidence interval at
each value of `lstat`, since it accounts for the irreducible error around the
regression line, not just the uncertainty in estimating the line itself.

## Visualizing the fit

```{r plot-fit}
plot(Boston$lstat, Boston$medv,
     xlab = "Lower status of the population (%)",
     ylab = "Median home value ($1000s)",
     main = "medv vs. lstat", pch = 20, col = "steelblue")
abline(lm.fit, lwd = 2, col = "firebrick")
```

Some extra styling options for `plot()` and `abline()`:

```{r plot-styles}
par(mfrow = c(2, 2))
plot(Boston$lstat, Boston$medv, col = "steelblue", pch = 20)
plot(Boston$lstat, Boston$medv, pch = "+")
plot(1:20, 1:20, pch = 1:20, main = "Point shapes 1-20")
plot(Boston$lstat, Boston$medv, pch = 20, cex = 0.6)
```

## Diagnostic plots

Base R's `plot()` applied directly to a fitted `lm` object produces four
standard diagnostic plots: residuals vs. fitted, a Q-Q plot, scale-location,
and residuals vs. leverage.

```{r diagnostics}
par(mfrow = c(2, 2))
plot(lm.fit)
```

We can also compute residuals and studentized residuals manually, and inspect
leverage statistics:

```{r residuals-leverage}
par(mfrow = c(1, 1))
plot(predict(lm.fit), residuals(lm.fit),
     xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, lty = 2)

plot(predict(lm.fit), rstudent(lm.fit),
     xlab = "Fitted values", ylab = "Studentized residuals")
abline(h = 0, lty = 2)

# Leverage statistics -- which observations have unusually high leverage?
plot(hatvalues(lm.fit),
     ylab = "Leverage", main = "Leverage by observation")
which.max(hatvalues(lm.fit))
```

# Multiple Linear Regression

A single predictor rarely tells the whole story — `medv` plausibly depends
on more than just `lstat`. Multiple linear regression extends the same
least-squares idea to $p$ predictors:

$$
\text{medv} = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_p X_p + \epsilon
$$

Each $\hat\beta_j$ is now interpreted as the effect of a one-unit increase
in $X_j$ *holding all other predictors fixed* — a different reading than
in simple regression, where $\hat\beta_1$ captured $X_1$'s effect with
nothing else in the model to hold constant.

Now we extend the model to include more than one predictor.

```{r mlr-two}
lm.fit2 <- lm(medv ~ lstat + age, data = Boston)
summary(lm.fit2)
```

## Using all predictors

Rather than typing every variable name, we can regress `medv` on all other
variables in the data frame using `.`:

```{r mlr-all}
lm.fit.all <- lm(medv ~ ., data = Boston)
summary(lm.fit.all)
```

We can access individual pieces of the summary object, such as $R^2$ and the
residual standard error:

```{r summary-pieces}
summary(lm.fit.all)$r.squared
summary(lm.fit.all)$sigma
```

## Checking for multicollinearity

Variance inflation factors (VIFs) help flag predictors that are highly
correlated with other predictors, which can inflate coefficient standard
errors. For predictor $X_j$, the VIF is defined using $R^2_{X_j}$ — the
$R^2$ from regressing $X_j$ on *all the other predictors*:

$$
\text{VIF}(\hat\beta_j) = \frac{1}{1 - R^2_{X_j}}
$$

If $X_j$ is well explained by the other predictors ($R^2_{X_j}$ close to 1),
the denominator shrinks toward 0 and the VIF blows up — a sign that $X_j$'s
own coefficient estimate is unstable because the data can't cleanly separate
its effect from the other predictors it's correlated with. A VIF of 1 means
no collinearity at all; VIFs above roughly 5–10 are typically treated as
cause for concern.

```{r vif}
vif(lm.fit.all)
```

## Fitting with one predictor excluded

Suppose we want to fit the model using all predictors except one (say, `age`,
which had a very high p-value above). We can do this two ways:

```{r drop-predictor}
lm.fit.no.age <- lm(medv ~ . - age, data = Boston)
summary(lm.fit.no.age)

# Equivalently, using update() on an existing fit:
lm.fit.updated <- update(lm.fit.all, ~ . - age)
summary(lm.fit.updated)
```

# Interaction Terms

Additive multiple regression assumes each predictor's effect is independent
of the others — the effect of `lstat` on `medv` is the same whichever value
`age` happens to take. An interaction term relaxes that assumption:

$$
\text{medv} = \beta_0 + \beta_1 \cdot \text{lstat} + \beta_2 \cdot \text{age} + \beta_3 \cdot (\text{lstat} \times \text{age}) + \epsilon
$$

Rearranging the `lstat` terms makes the point concrete — `lstat`'s effective
slope now depends on the value of `age`:

$$
\text{medv} = \beta_0 + (\beta_1 + \beta_3 \cdot \text{age}) \cdot \text{lstat} + \beta_2 \cdot \text{age} + \epsilon
$$

`lstat:age` includes only the interaction term ($\beta_3$), while `lstat*age`
is shorthand for `lstat + age + lstat:age` -- i.e., it includes both main
effects and the interaction, matching the full formula above.

```{r interactions}
summary(lm(medv ~ lstat * age, data = Boston))
```

# Non-linear Transformations of Predictors

The `lm()` function can accommodate non-linear transformations of predictors
directly in the formula, using `I()` to protect arithmetic operators like `^`
from being interpreted as formula syntax.

$$
\text{medv} = \beta_0 + \beta_1 \cdot \text{lstat} + \beta_2 \cdot \text{lstat}^2 + \epsilon
$$

This is still a *linear* model in the statistical sense — linear in the
parameters $\beta_0, \beta_1, \beta_2$ — even though the fitted curve itself
is not a straight line. That's exactly why `lm()` can fit it directly.

```{r poly-term}
lm.fit.sq <- lm(medv ~ lstat + I(lstat^2), data = Boston)
summary(lm.fit.sq)
```

We can use `anova()` to formally compare the linear-only model to the model
that also includes the quadratic term:

```{r anova-compare}
lm.fit.linear <- lm(medv ~ lstat, data = Boston)
anova(lm.fit.linear, lm.fit.sq)
```

```{r poly-diagnostics}
par(mfrow = c(2, 2))
plot(lm.fit.sq)
```

For a higher-order polynomial fit, the `poly()` function is more convenient
than writing out each `I(lstat^k)` term by hand:

```{r poly-function}
lm.fit.poly5 <- lm(medv ~ poly(lstat, 5), data = Boston)
summary(lm.fit.poly5)
```

We can also try a log transformation of a predictor:

```{r log-term}
summary(lm(medv ~ log(rm), data = Boston))
```

# Qualitative Predictors

We switch to the `Carseats` data set, which includes a mix of quantitative
and qualitative (categorical) predictors, to demonstrate how `lm()` handles
factor variables.

```{r carseats-explore}
Carseats <- ISLR2::Carseats
str(Carseats)
```

## Fitting a model with interactions and a factor variable

```{r carseats-fit}
lm.fit.carseats <- lm(Sales ~ . + Income:Advertising + Price:Age,
                       data = Carseats)
summary(lm.fit.carseats)
```

R automatically creates dummy variables for qualitative predictors such as
`ShelveLoc` (a three-level factor: `Bad`, `Medium`, `Good`). With $k=3$
levels, R creates $k-1=2$ dummy variables and folds their coefficients into
the same linear model:

$$
\text{Sales} = \beta_0 + \beta_1 \cdot X_{\text{Good}} + \beta_2 \cdot X_{\text{Medium}} + \cdots + \epsilon,
\qquad X_{\text{Good}} =
\begin{cases} 1 & \text{if ShelveLoc = Good} \\ 0 & \text{otherwise} \end{cases}
$$

(and similarly for $X_{\text{Medium}}$). The omitted level, `Bad`, becomes
the baseline: when both dummies are 0, the model's intercept $\beta_0$
represents a `Bad` shelf location, so $\beta_1$ and $\beta_2$ are read as
the shift in expected `Sales` *relative to* `Bad`, not as standalone effects.
We can check how these dummy variables were coded using `contrasts()`:

```{r contrasts}
contrasts(Carseats$ShelveLoc)
```

The coefficients on `ShelveLocGood` and `ShelveLocMedium` represent the
expected change in sales relative to the baseline level (`Bad` shelf
location), holding all other predictors fixed.

# Writing a Small Function

Finally, a short example of writing our own function -- here, one that loads
several packages at once, which can be handy at the start of a script or
report like this one.

```{r custom-function}
LoadLibraries <- function() {
  library(ISLR2)
  library(MASS)
  library(car)
  print("The libraries have been loaded.")
}

LoadLibraries()
```

# Summary

In this lab we:

1. Fit and interpreted simple and multiple linear regression models.
2. Computed confidence and prediction intervals for predicted responses.
3. Produced and interpreted standard regression diagnostic plots.
4. Checked for multicollinearity using variance inflation factors.
5. Added interaction terms and non-linear (polynomial, log) transformations.
6. Incorporated qualitative predictors and inspected their dummy-variable
   coding.

These tools form the foundation for the more flexible modeling approaches
covered in later chapters.
