---
fontsize: 11pt
geometry: margin=1in
output:
  pdf_document:
    keep_tex: true
    number_sections: true
    extra_dependencies: ["amsmath"]
header-includes:
  - \input{../../_common/preamble.tex}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 4
)
```

```{=latex}
\makecoverpage{Chapter 3 --- Linear Regression \\ Exercises 1, 6, 11}
```

\tableofcontents
\newpage

# Conceptual Exercises

## Exercise 1 --- Null Hypotheses in Table 3.4

**Question.** Describe the null hypotheses to which the p-values given in Table 3.4
correspond. Explain what conclusions you can draw based on these p-values.
Your explanation should be phrased in terms of `sales`, `TV`, `radio`, and
`newspaper`, rather than in terms of the coefficients of the linear model.

**Setting up the model.** The Advertising data are fit with a single multiple
regression model in which `sales` is regressed simultaneously on all three
advertising budgets:

$$
\text{sales} = \beta_0 + \beta_1 \cdot \text{TV} + \beta_2 \cdot \text{radio} + \beta_3 \cdot \text{newspaper} + \epsilon .
$$

Table 3.4 (reproduced below from the text) reports the estimated coefficient,
its standard error, the resulting $t$-statistic, and the p-value for each
term:

| Predictor   | Coefficient | Std. error | $t$-statistic | p-value  |
|:------------|------------:|-----------:|---------------:|:---------|
| Intercept   | 2.939       | 0.3119     | 9.42           | < 0.0001 |
| TV          | 0.046       | 0.0014     | 32.81          | < 0.0001 |
| radio       | 0.189       | 0.0086     | 21.89          | < 0.0001 |
| newspaper   | $-0.001$    | 0.0059     | $-0.18$        | 0.8599   |

**Plain-language explanation.** Each p-value in the table answers one
narrow question: *if I already know how much was spent on the other two
advertising channels, does knowing the spend on this particular channel
still help me predict `sales`?* This is different from asking whether the
channel is correlated with `sales` on its own — it is asking whether the
channel carries **extra**, non-redundant information about `sales` once the
other two channels are held fixed.

- For **TV**, the p-value is tiny (`< 0.0001`). This says: it is essentially
  impossible that TV spending has *no* effect on `sales` once radio and
  newspaper spending are held constant — the data would be extremely
  surprising under that assumption. So we conclude TV advertising is
  genuinely associated with higher sales, over and above whatever radio and
  newspaper are doing.
- For **radio**, the same reasoning applies: the p-value is tiny, so radio
  spending appears to have a real, additional effect on sales, holding TV
  and newspaper fixed.
- For **newspaper**, the p-value is large (`0.8599`). This says the data are
  entirely consistent with newspaper spending having *no* additional effect
  on sales once TV and radio are already accounted for. We do **not** have
  evidence that spending more on newspaper advertising increases sales,
  after controlling for the other two channels.

**Formal statement.** For each coefficient $\beta_j$, $j \in \{\text{TV},
\text{radio}, \text{newspaper}\}$, the p-value corresponds to testing

$$
H_0 : \beta_j = 0 \qquad \text{vs.} \qquad H_a : \beta_j \neq 0,
$$

holding the other two predictors fixed. At any conventional significance
level (e.g. $\alpha = 0.05$), we **reject** $H_0$ for TV and radio (strong
evidence of association with `sales`), and we **fail to reject** $H_0$ for
newspaper (no evidence of association with `sales`, once TV and radio are
already in the model).

**A subtlety worth flagging.** This does *not* mean newspaper spending is
unrelated to sales in a simple, one-predictor regression — on its own it is
mildly positively correlated with `sales`. The reason it drops out here is
that newspaper spending is itself correlated with radio spending (markets
that spend more on radio also tend to spend more on newspaper). Once radio
is in the model, it "absorbs" the association that newspaper appeared to
have on its own, and newspaper has nothing left to contribute. This is a
classic confounding pattern, not a contradiction.

\newpage

## Exercise 6 --- The Least Squares Line Passes Through $(\bar x,\bar y)$

**Question.** Using (3.4), argue that in the case of simple linear
regression, the least squares line always passes through the point
$(\bar x, \bar y)$.

**Plain-language idea.** The least squares line is chosen to minimize the
total squared vertical distance between the data and the line. It turns out
that the intercept term is *defined* in exactly the way needed to force the
line through the "average point" of the data — the point whose coordinates
are the mean of $x$ and the mean of $y$. So this isn't a coincidence that
happens to hold for some datasets; it's built into the formula for
$\hat\beta_0$ itself, and therefore holds for *every* dataset.

**Formal argument.** Equation (3.4) gives the least squares estimates for
simple linear regression as

$$
\hat\beta_1 = \frac{\sum_{i=1}^n (x_i - \bar x)(y_i - \bar y)}{\sum_{i=1}^n (x_i - \bar x)^2},
\qquad
\hat\beta_0 = \bar y - \hat\beta_1 \bar x .
$$

The fitted regression line is $\hat y = \hat\beta_0 + \hat\beta_1 x$.
Evaluate this line at $x = \bar x$:

$$
\hat y \Big|_{x = \bar x} \;=\; \hat\beta_0 + \hat\beta_1 \bar x
\;=\; \big(\bar y - \hat\beta_1 \bar x\big) + \hat\beta_1 \bar x
\;=\; \bar y .
$$

The $\hat\beta_1 \bar x$ terms cancel exactly, leaving $\hat y = \bar y$ when
$x = \bar x$. Hence the point $(\bar x, \bar y)$ always satisfies the fitted
equation — that is, it always lies on the least squares line, regardless of
the particular data values. $\blacksquare$

**Numerical check.** The algebra above holds for *any* numeric data, so the
following chunk checks it on one arbitrary, fixed data set. Because the
data are fixed constants (no random number generation involved), the result
is fully deterministic:

```{r ex6-demo}
x <- c(1, 2, 3, 4, 5, 6, 7)
y <- c(2.1, 3.4, 5.0, 4.8, 6.3, 7.1, 7.9)

fit <- lm(y ~ x)

xbar <- mean(x)
ybar <- mean(y)
y_hat_at_xbar <- as.numeric(predict(fit, newdata = data.frame(x = xbar)))

data.frame(
  quantity = c("mean(x)", "mean(y)", "fitted value at mean(x)"),
  value    = c(xbar, ybar, y_hat_at_xbar)
)

isTRUE(all.equal(y_hat_at_xbar, ybar))
```

The final line evaluates to `TRUE`: the fitted value at $\bar x$ equals
$\bar y$ exactly (up to floating-point rounding), confirming the proof
above.

\newpage

# Applied Exercises

## Exercise 11 --- Regression Through the Origin

**Question setup.** We generate a predictor `x` and a response `y` as
follows, and investigate the $t$-statistic for $H_0: \beta = 0$ in simple
linear regression **without an intercept**.

```{r ex11-setup}
set.seed(1)
x <- rnorm(100)
y <- 2 * x + rnorm(100)
```

### (a) Regress `y` onto `x`, without an intercept

```{r ex11a}
fit_yx <- lm(y ~ x + 0)
summary(fit_yx)
```

**Comment.** Since `y` was generated as (approximately) $2x$ plus noise, we
would expect the slope estimate $\hat\beta$ to come out close to $2$, with a
standard error that is small relative to the estimate, and hence a large
$t$-statistic and a very small p-value — strong evidence against
$H_0 : \beta = 0$. Compare these expectations against the actual `Estimate`,
`Std. Error`, `t value`, and `Pr(>|t|)` printed in the `summary()` output
above: the fitted slope should sit close to $2$, and the association should
come through as highly statistically significant.

### (b) Regress `x` onto `y`, without an intercept

```{r ex11b}
fit_xy <- lm(x ~ y + 0)
summary(fit_xy)
```

**Comment.** Now the roles are reversed: we predict `x` from `y`. Since
$y \approx 2x$, inverting the relationship gives $x \approx y/2$, so we
would expect the slope estimate here to be roughly $0.5$ — noticeably
different from the estimate in part (a). What is more interesting is what
does *not* change: as parts (c)–(e) show, the $t$-statistic testing
$H_0 : \beta = 0$ is identical in both directions, even though the
coefficient estimates and their standard errors differ.

### (c) Relationship between (a) and (b)

Both regressions are, in effect, testing the same underlying question:
*is there a linear association between `x` and `y`?* The coefficient
estimates in (a) and (b) are not reciprocals of one another and have
different standard errors, because they are measured on different scales
(slope of $y$ on $x$ versus slope of $x$ on $y$). However, the strength of
evidence for association — captured by the $t$-statistic and p-value — turns
out to be exactly the same in both directions. Parts (d) and (e) show why
this must be true algebraically, and the R output above/below confirms it
numerically.

### (d) Algebraic derivation of the $t$-statistic formula

For regression through the origin, (3.38) gives

$$
\hat\beta = \frac{\sum_{i=1}^n x_i y_i}{\sum_{i'=1}^n x_{i'}^2},
$$

and the exercise gives the standard error as

$$
SE(\hat\beta) = \sqrt{\dfrac{\sum_{i=1}^n (y_i - x_i \hat\beta)^2}{(n-1)\sum_{i'=1}^n x_{i'}^2}} .
$$

We want to show that $t = \hat\beta / SE(\hat\beta)$ can be rewritten as

$$
t = \frac{\sqrt{n-1}\,\sum_{i=1}^n x_i y_i}{\sqrt{\left(\sum_{i=1}^n x_i^2\right)\left(\sum_{i'=1}^n y_{i'}^2\right) - \left(\sum_{i'=1}^n x_{i'} y_{i'}\right)^2}} .
$$

**Step 1 — expand the residual sum of squares.** Write
$S_{xy} = \sum_i x_i y_i$, $S_{xx} = \sum_i x_i^2$, $S_{yy} = \sum_i y_i^2$.
Expanding the square:

$$
\sum_i (y_i - x_i\hat\beta)^2 = \sum_i y_i^2 - 2\hat\beta \sum_i x_i y_i + \hat\beta^2 \sum_i x_i^2
= S_{yy} - 2\hat\beta S_{xy} + \hat\beta^2 S_{xx}.
$$

**Step 2 — substitute $\hat\beta = S_{xy}/S_{xx}$.**

$$
2\hat\beta S_{xy} = \frac{2 S_{xy}^2}{S_{xx}}, \qquad
\hat\beta^2 S_{xx} = \frac{S_{xy}^2}{S_{xx}} .
$$

So

$$
\sum_i (y_i - x_i\hat\beta)^2 = S_{yy} - \frac{2S_{xy}^2}{S_{xx}} + \frac{S_{xy}^2}{S_{xx}}
= S_{yy} - \frac{S_{xy}^2}{S_{xx}}
= \frac{S_{xx}S_{yy} - S_{xy}^2}{S_{xx}} .
$$

**Step 3 — assemble the $t$-statistic.**

$$
t = \frac{\hat\beta}{SE(\hat\beta)}
= \frac{S_{xy}/S_{xx}}{\sqrt{\dfrac{1}{(n-1)S_{xx}}\left(\dfrac{S_{xx}S_{yy}-S_{xy}^2}{S_{xx}}\right)}}
= \frac{S_{xy}/S_{xx}}{\sqrt{\dfrac{S_{xx}S_{yy}-S_{xy}^2}{(n-1)S_{xx}^2}}} .
$$

Since $\sqrt{1/S_{xx}^2} = 1/S_{xx}$ (as $S_{xx} > 0$), the $S_{xx}$ in the
denominator of $\hat\beta$ cancels with the $S_{xx}$ pulled out of the
square root:

$$
t = \frac{S_{xy}/S_{xx}}{\dfrac{1}{S_{xx}}\sqrt{\dfrac{S_{xx}S_{yy}-S_{xy}^2}{n-1}}}
= \frac{S_{xy}}{\sqrt{\dfrac{S_{xx}S_{yy}-S_{xy}^2}{n-1}}}
= \frac{\sqrt{n-1}\; S_{xy}}{\sqrt{S_{xx}S_{yy}-S_{xy}^2}},
$$

which is exactly the target formula, written back in full as

$$
t = \frac{\sqrt{n-1}\sum_i x_iy_i}{\sqrt{\left(\sum_i x_i^2\right)\left(\sum_i y_i^2\right) - \left(\sum_i x_iy_i\right)^2}}. \qquad \blacksquare
$$

**Numerical confirmation in R.** The chunk below computes $t$ three
independent ways — by hand from $\hat\beta$ and $SE(\hat\beta)$, from the
closed-form formula just derived, and by reading it off `lm()`'s own
summary — and checks that all three agree:

```{r ex11d}
n <- length(x)

beta_hat <- sum(x * y) / sum(x^2)
se_beta  <- sqrt(sum((y - x * beta_hat)^2) / ((n - 1) * sum(x^2)))
t_manual <- beta_hat / se_beta

t_formula <- (sqrt(n - 1) * sum(x * y)) /
  sqrt(sum(x^2) * sum(y^2) - (sum(x * y))^2)

t_lm <- summary(fit_yx)$coefficients["x", "t value"]

data.frame(
  method = c("by hand (beta_hat / SE)", "closed-form formula", "lm() summary"),
  t_stat = c(t_manual, t_formula, as.numeric(t_lm))
)

isTRUE(all.equal(t_manual, t_formula)) &&
  isTRUE(all.equal(t_manual, as.numeric(t_lm)))
```

The final line should evaluate to `TRUE`, confirming that the algebra above
matches both a from-scratch numerical computation and what `lm()` reports.

### (e) The $t$-statistic is the same for `y` on `x` as for `x` on `y`

Look at the closed-form expression from part (d):

$$
t = \frac{\sqrt{n-1}\sum_i x_i y_i}{\sqrt{\left(\sum_i x_i^2\right)\left(\sum_i y_i^2\right) - \left(\sum_i x_i y_i\right)^2}} .
$$

This formula treats $x$ and $y$ almost, but not quite, symmetrically — the
only way $x$ and $y$ enter is through the three summary quantities
$S_{xy} = \sum_i x_iy_i$, $S_{xx} = \sum_i x_i^2$, and $S_{yy} = \sum_i
y_i^2$. If instead we regress $x$ onto $y$ (without an intercept), the
same derivation applies with the roles of $x$ and $y$ swapped throughout,
giving

$$
t' = \frac{\sqrt{n-1}\sum_i y_i x_i}{\sqrt{\left(\sum_i y_i^2\right)\left(\sum_i x_i^2\right) - \left(\sum_i y_ix_i\right)^2}} .
$$

But $\sum_i x_iy_i = \sum_i y_ix_i$ (multiplication of real numbers is
commutative), and the product $\left(\sum_i y_i^2\right)\left(\sum_i
x_i^2\right)$ is the same regardless of the order in which the two sums are
multiplied. So the numerator and denominator of $t'$ are literally the same
expressions as in $t$, just written in a different order:
$t' = t$. This is exactly the pattern already glimpsed numerically in parts
(a)–(b): the coefficient estimates differ, but the $t$-statistic for
$H_0 : \beta = 0$ does not.

### (f) With an intercept: same check in R

```{r ex11f}
fit_yx_int <- lm(y ~ x)
fit_xy_int <- lm(x ~ y)

t_yx_int <- summary(fit_yx_int)$coefficients["x", "t value"]
t_xy_int <- summary(fit_xy_int)$coefficients["y", "t value"]

data.frame(
  regression = c("y ~ x", "x ~ y"),
  t_stat     = c(as.numeric(t_yx_int), as.numeric(t_xy_int))
)

isTRUE(all.equal(as.numeric(t_yx_int), as.numeric(t_xy_int)))
```

The final line should again evaluate to `TRUE`. This is not a coincidence
special to regression through the origin: for simple linear regression
**with** an intercept, it is a standard result that the $t$-statistic for
testing $H_0 : \beta_1 = 0$ satisfies

$$
t^2 = \frac{(n-2)\, r_{xy}^2}{1 - r_{xy}^2},
$$

where $r_{xy}$ is the sample correlation between $x$ and $y$. Because
correlation is symmetric ($r_{xy} = r_{yx}$), this expression is identical
whether we regress $y$ on $x$ or $x$ on $y$ — so the two regressions must
again produce the same $t$-statistic, exactly as the R output above
confirms.

\newpage

